---
title: "Ashfield 2019-2023 spatial analyses"
output: 
  bookdown::html_document2:
    fig_width: 6.4
    fig_height: 4.8
    number_sections: no
    self_contained: no
    df_print: default
    fig_caption: yes
    smart: no
---

```{css, echo=FALSE}
table {
  margin: auto;
  border-top: 1px solid #333;
  border-bottom: 1px solid #333;
}
table thead th { border-bottom: 1px solid #ddd; }
th, td { padding: 5px; }
thead, tfoot, tr:nth-child(even) { background: #eee; }
.striped tr:nth-child(even) { background: #eee; }
```

```{r load packages and environment but do not show, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}
library(sf)
library(lctools)
library(gstat)
library(maptiles)
library(prettymapr)
library(fields)
library(viridis)
library(scico)
library(TeachingDemos)
library(ncf)
library(flextable)
  set_flextable_defaults(font.family = "sans", theme_fun = "theme_zebra", 
        font.size = 11, text.align = "right",padding.left = 1, padding.right = 1,
        padding.top = 2, padding.bottom = 2)
knitr::opts_chunk$set(comment="")
```

```{r importing the data, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
git <- "https://raw.githubusercontent.com/Ratey-AtUWA/Ashfield-REE/main/"
afs1923 <- read.csv(paste0(git,"afs1923.csv"), stringsAsFactors = TRUE)
afs1923$Year <- as.factor(afs1923$Year)
af_border <- read.csv(paste0(git,"afr_hull.csv"))
afs1923surf <- afs1923[-which(afs1923$Depth_upper_cm > 1),]
afs1923surf[which(afs1923surf$Type=="Soil"),"Type"] <- 
  rep("Saltmarsh",length(which(afs1923surf$Type=="Soil")))
afs1923surf <- droplevels(afs1923surf)
afs1923surf$Type <- factor(afs1923surf$Type, 
                 levels = c("Drain_Sed", "Lake_Sed", "Saltmarsh", "Other"))
afs1923utm <- st_as_sf(afs1923surf, coords = c("Easting","Northing"), 
                       crs = st_crs(32750))
afs1923ll <- st_transform(afs1923utm, crs = st_crs(4326))
afr_map <- read.csv(paste0(git,"afr_map_v3.csv"))
```

# Making our base map of Ashfield Flats

### Define coordinate reference systems

```{r define-CRS, warning=FALSE, error=FALSE, results='hold', echo=FALSE}
LongLat <- st_crs(4326) # uses Earth ellipsis specs from WGS84 datum
UTM50S <- st_crs(32750) # just for Zone 50, S hemisphere!
```

### ├ Map extent object

```{r make-map-extent-object, warning=FALSE, error=FALSE, results='hold', echo=FALSE}
extent <- st_as_sf(x = data.frame(x = c(399860,400580),
                                  y = c(6467870,6468350)),
                   coords = c("x", "y"), crs = UTM50S)
st_coordinates(extent)
```

### ├ Getting and plotting the map tile data

```{r basemap, fig.height=5.75, fig.width=8, fig.align='center', fig.cap="Map of Ashfield Flats Reserve and adjacent area (UTM projection, Zone 50S, EPSG:32750) used subsequently as the base map for spatial analyses. Generated using the maptiles R package, with OpenStreetMap tiles.", out.width="75%", warning=FALSE, message=FALSE, results='hide', echo=FALSE}
afstiles <- get_tiles(extent, provider="Thunderforest.OpenCycleMap",
             apikey="03b0375e8e7c47d0a03018a9001de439",
                      crop = TRUE, zoom = 17) # make map object

par(oma=c(3,3,1,1), mar=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), 
    lend=2, ljoin=1, tcl=0.3)
plot_tiles(afstiles)
axis(1)
mtext(side=1, line=2, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.2)
axis(2)
mtext(side=2, line=2, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
with(afr_map, polygon(wetland_E, wetland_N, border = "steelblue", col="#87CEEB80"))
with(afs1923surf, points(Easting, Northing, pch = 3, cex = 0.75, col = "navy"))
with(af_border, lines(Easting,Northing, lty = 3, col = "darkgreen"))
addnortharrow()
addscalebar(plotepsg = 32750, pos = "bottomleft")
box()
```

#### **Note**: in this document REE (&equiv; &sum;REE) = Ce + La + Nd + Gd.
&sum;REE does not include yttrium (Y) at present, but it could... using <br>
`afs1923$YREE <- afs1923$REE + afs1923$Y`.

## Spatial Autocorrelation

### ├ Calculate Global Moran's I

```{r global-Moran, echo=TRUE, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
var0 <- "REE" # choose the variable of interest

# Calculate global Moran's I
data_temp <- na.omit(afs1923surf[,c("Easting", "Northing", var0)])
Coords <- cbind(data_temp$Easting, data_temp$Northing)
bw <- 8 # bw = bandwidth, = number of nearest point for calculation
mI <- moransI(Coords, bw, data_temp[,3])
cat("Global Moran's I for ",var0,"; from Ashfield Flats 2019-2023\n", sep="");
t(as.matrix(mI[c(2,4:7)]))
rm(list = c("data_temp","Coords","bw","mI","var0")) # remove temporary objects
```

### ├ Plot local Moran's I

```{r local-Moran, message=FALSE, warning=FALSE, fig.height=5.75, fig.width=8, fig.align='center', out.width="80%", fig.cap="Map of Local Moran's I for REE concentrations in Ashfield Flats sediments 2019-2023. The Global Moran's I parameter is also shown beneath the legend.", results='hide', echo=FALSE}
var0 <- "REE" # choose the variable of interest
data_temp <- na.omit(afs1923surf[,c("Easting", "Northing", var0)])
Coords <- cbind(data_temp$Easting, data_temp$Northing)
mI <- moransI(Coords, 8, log10(data_temp[,3])) # log10 minimises skewness
local_moran <- l.moransI(Coords, 8, log10(data_temp[,3]), scatter.plot = FALSE)
plotdata <- data.frame(Easting=Coords[,1], Northing=Coords[,2], 
                       MoranI=local_moran$Ii, p_value=local_moran$p.value)
pos0 <- subset(plotdata, plotdata$MoranI>0 & plotdata$p_value<=0.05)
neg0 <- subset(plotdata, plotdata$MoranI<0 & plotdata$p_value<=0.05)
#
par(oma=c(3,3,1,1), mar=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), 
    lend=2, ljoin=1, tcl=0.3, tcl = 0.5, lwd = 1)
plot_tiles(afstiles)
with(afr_map, polygon(wetland_E, wetland_N, border = "steelblue", col="#87CEEB80"))
axis(1)
mtext(side=1, line=2, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.2)
axis(2)
mtext(side=2, line=2, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
addnortharrow(pos="topright", border=1, lwd=1, text.col=1,  
              padin=c(0.1,0.2), scale=1.2)
addscalebar(plotepsg = 32750)
box()
points(Coords, pch=3, cex=0.5, col = "grey67")
with(pos0, symbols(Easting, Northing, circles = 15*sqrt(MoranI*0.02), 
                   inches = F, fg = "blue3", bg = "#00008080", add= TRUE))
with(neg0, symbols(Easting, Northing, squares = 15*sqrt(MoranI*-0.04), 
                   inches = F, fg = "red3", bg = "#80000080", add= TRUE))
# rect(107,-18,120,-12.5, col="#E4DDB7", border = "white", lwd = 2)
legend("bottomright", bty = "o", cex = 1, title.cex = 0.9,
       legend = c("Positive I", "Negative I",
                  "", NA),
       title = paste0("Local Moran's I for log(",var0, 
                      ") in afs1923\n(p \u2264 0.05 only)"),
       pch = c(21, 22,NA,NA), pt.cex = c(2, 1.7,NA,NA),
       col = c("blue3", "red3"), pt.bg = c("#00008080", "#80000080"),
       title.col = "navy", y.intersp = 0.9)
text(400600,6467860, 
     labels = paste("Global Moran's I =", signif(as.numeric(mI[2]),3),
                               "\nRandomization p =", signif(as.numeric(mI[7]),3)), 
     pos = 2)
```

### ├ Plot 'LISA'

```{r lisa-plot, message=FALSE, warning=FALSE, fig.height=5.75, fig.width=8, fig.align='center', out.width="80%", fig.cap="Map of LISA for REE concentrations in the Ashfield Flats sediment data, 2019-2023.", results='hide',echo=FALSE}
data_temp <- na.omit(as.data.frame(cbind(afs1923surf$Easting,afs1923surf$Northing,
                                         afs1923surf$REE)))
colnames(data_temp) <- c("Easting","Northing","REE")


Coords <- cbind(data_temp$Easting, data_temp$Northing)
bw <- 8
locMI <- l.moransI(Coords,bw,data_temp$REE, scatter.plot = F)
LISA <- as.data.frame(cbind(Coords[,1:2], log10(data_temp$REE), 
                            locMI$Ii, 
                            locMI$p.value))
colnames(LISA) <- c("Easting", "Northing", "REE", "MoranI", "p_value") # ;head(LISA)
medREE <- median(LISA$REE, na.rm = TRUE)
HiHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA$REE>=medREE)
LoLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA$REE<medREE)
HiLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA$REE>=medREE)
LoHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA$REE<medREE)
#
par(oma=c(3,3,1,1), mar=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), 
    lend=2, ljoin=1, tcl=0.3, tcl = 0.5, lwd = 1)
palette(c("black","#400000","blue3","orange","skyblue",
          "#FF8080","#8080FF","#FFA50040","#87CEEB40","white"))
plot_tiles(afstiles)
with(afr_map, polygon(wetland_E, wetland_N, border = "steelblue", col="#87CEEB80"))
axis(1)
mtext(side=1, line=1.6, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.2)
axis(2)
mtext(side=2, line=1.6, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
addnortharrow(pos="topright", border=1, lwd=1, text.col=1,  
              padin=c(0.1,0.2), scale=1.2)
addscalebar(plotepsg = 32750)
box()
with(data_temp, points(Easting, Northing, pch=3, cex = 0.7, col="grey"))
with(HiHi, points(Northing ~ Easting, pch = 22,
                  col = 2, bg = 6, lwd = 2, cex = 1.8))
with(LoLo, points(Northing ~ Easting, pch = 21, 
                  col = 3, bg = 7, lwd = 2, cex = 2))
with(HiLo, points(Northing ~ Easting, pch = 25,  
                  col = 2, bg = 7, lwd = 3, cex = 1.4))
with(LoHi, points(Northing ~ Easting, pch = 24,
                  col = 3, bg = 6, lwd = 3, cex = 1.4))
legend("bottomright", bty = "o", inset = 0.015, cex = 1, ncol = 3,
       legend = c("High-High", "Low-Low","High-Low","Low-High","Sample points"),
       pch = c(22, 21, 25,24,3), pt.cex = c(1.8,2,1.4,1.4,0.7), 
       pt.lwd = c(2,2,3,3,1),
       col = c(2,3,2,3,"grey"), pt.bg = c(6,7,7,6,NA), text.col = 1,
       box.col = "#b0b0b0", box.lwd = 1.3, bg = "#E8E8E8", 
       title="\nSpatial autocorrelation classes for REE",
       y.intersp = 1.1)
```

### ├ LISA maps for multiple variables

```{r multi-LISA, fig.height=15.96, fig.width=12, fig.align='center', fig.cap="LISA autocorrelation maps for REE, Al, Fe, S, As, Cu, Pb, and Zn concentrations in the Ashfield Flats sediment data, 2019-2023.", message=FALSE, warning=FALSE, echo=FALSE}
par(mfrow=c(4,2), mar=c(0,0,0,0), oma=c(3,3,0.5,1.5), mgp=c(1.6,0.2,0), 
    tcl=0.3,lend="square", ljoin="mitre")
palette(c("black","#400000","blue3","orange","skyblue",
          "#FF8080","#8080FF","#FFA50040","#87CEEB40","white"))
varz <- c("REE","Al","Fe","S","As","Cu","Pb","Zn")
for (i in 1:length(varz)){
  data_temp <- na.omit(as.data.frame(cbind(afs1923$Easting,afs1923$Northing,
                                         afs1923[,varz[i]])))
  colnames(data_temp) <- c("Easting","Northing",varz[i])

  Coords <- cbind(data_temp$Easting, data_temp$Northing)
  bw <- 8
  locMI <- l.moransI(Coords,bw,data_temp[,varz[i]], scatter.plot = F)
  LISA <- as.data.frame(cbind(Coords[,1:2], log10(data_temp[,varz[i]]), 
                            locMI$Ii, 
                            locMI$p.value))
  colnames(LISA) <- c("Easting", "Northing", var0, "MoranI", "p_value")
  medREE <- median(LISA[,var0], na.rm = TRUE)
  HiHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA[,var0]>=medREE)
  LoLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA[,var0]<medREE)
  HiLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA[,var0]>=medREE)
  LoHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA[,var0]<medREE)

  plot_tiles(afstiles)
  with(afr_map, polygon(wetland_E, wetland_N, border = "steelblue", col="#87CEEB80"))
  if(i > 6.5) {
    axis(1, at = seq(399900,400500,100), 
     labels = c("399900","400000","400100","400200","400300","400400","400500"))
    mtext(side=1, line=1.6, text="Easting (UTM Zone 50, m)", font=2, cex=1)
    }
  if(round(i/2,0) != (i/2)) {
    axis(2)
    mtext(side=2, line=1.6, text="Northing (UTM Zone 50, m)", font=2, cex=1)
  }
  axis(1, labels=F); axis(2, labels = F)
  addnortharrow(pos="topleft", border=1, lwd=1, text.col=1,  
              padin=c(0.08,0.08), scale = 0.7)
  addscalebar(plotepsg = 32750, label.cex = 1.6)
  box()
  with(data_temp, points(Easting, Northing, pch=3, cex = 0.7, col="grey"))
  with(HiHi, points(Northing ~ Easting, pch = 22,
                  col = 2, bg = 6, lwd = 2, cex = 1.8))
  with(LoLo, points(Northing ~ Easting, pch = 21, 
                  col = 3, bg = 7, lwd = 2, cex = 2))
  with(HiLo, points(Northing ~ Easting, pch = 25,  
                  col = 2, bg = 7, lwd = 3, cex = 1.4))
  with(LoHi, points(Northing ~ Easting, pch = 24,
                  col = 3, bg = 6, lwd = 3, cex = 1.4))
  mtext(varz[i],3,-2,adj=0.98,font=2,cex=1.5)
}
rect(400090,6467855,400590,6467930,col="#FFFFFF80", border = "#B0B0B080")
legend("bottomright", bty = "n", cex = 1.6, ncol=3,
       legend = c("High-High", "Low-Low","High-Low","Low-High","Samples"),
       pch = c(22, 21, 25,24,3), pt.cex = c(1.8,2,1.4,1.4,0.7), 
       pt.lwd = c(2,2,2,2,1),
       col = c(2,3,2,3,"grey"), pt.bg = c(6,7,7,6,NA), text.col = 1,
       box.col = "#b0b0b0", box.lwd = 1.3, bg = "#E8E8E8",
       x.intersp = 0.8, inset=c(0.04,0))
text(400340,6467935,pos=3,labels="Spatial autocorrelation key", font=3,
     cex=1.6, offset = 0.15)
par(mfrow=c(1,1), mar=c(3,3,0.5,0.5), oma=c(0,0,0,0), mgp=c(1.6,0.2,0), 
    tcl=0.3,lend="square", ljoin="mitre")

```

### ├ LISA maps for REE, Al only

```{r REE-Al-LISA, fig.height=6.6, fig.width=18, fig.align='center', fig.cap="LISA autocorrelation maps for (a) ∑REE and (b) Al concentrations in the Ashfield Flats sediment data, 2019-2023.", message=FALSE, warning=FALSE, echo=FALSE}
par(mfrow=c(1,2), mar=c(0,0,0,0), oma=c(3,3,0.5,1.5), mgp=c(1.6,0.2,0), 
    tcl=0.3,lend="square", ljoin="mitre")
palette(c("black","#402000","blue3","#a58400","skyblue",
                 "#ffd000","#8080FF","#FFA50040","#87CEEB40","white"))
varz <- c("REE","Al") 
for (i in 1:length(varz)){
  data_temp <- na.omit(as.data.frame(cbind(afs1923$Easting,afs1923$Northing,
                                         afs1923[,varz[i]])))
  colnames(data_temp) <- c("Easting","Northing",varz[i])

  Coords <- cbind(data_temp$Easting, data_temp$Northing)
  bw <- 8
  locMI <- l.moransI(Coords,bw,data_temp[,varz[i]], scatter.plot = F)
  LISA <- as.data.frame(cbind(Coords[,1:2], log10(data_temp[,varz[i]]), 
                            locMI$Ii, 
                            locMI$p.value))
  colnames(LISA) <- c("Easting", "Northing", var0, "MoranI", "p_value")
  medREE <- median(LISA[,var0], na.rm = TRUE)
  HiHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA[,var0]>=medREE)
  LoLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA[,var0]<medREE)
  HiLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA[,var0]>=medREE)
  LoHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA[,var0]<medREE)

  plot_tiles(afstiles)
  with(afr_map, polygon(wetland_E, wetland_N, border = "steelblue", col="#87CEEB80"))
  axis(1, cex.axis=1.4, at = seq(399900,400600,100), 
       labels = c("399900","400000","400100","400200","400300","400400","400500",""))
  mtext(side=1, line=1.6, text="Easting (UTM Zone 50, m)", font=2, cex=1.6)
  if(round(i/2,0) != (i/2)) {
    axis(2, cex.axis=1.4)
    mtext(side=2, line=1.6, text="Northing (UTM Zone 50, m)", font=2, cex=1.6)
  }
  addnortharrow(pos="topleft", border=1, lwd=1, text.col=1,  
              padin=c(0.08,0.08), scale = 1.25)
  addscalebar(plotepsg = 32750, label.cex = 1.8, htin=0.2, widthhint=0.15)
  box()
  with(data_temp, points(Easting, Northing, pch=3, cex = 0.85, col="grey65"))
  with(HiHi, points(Northing ~ Easting, pch = 22,
                    col = 2, bg = 6, lwd = 2, cex = 1.8))
  with(LoLo, points(Northing ~ Easting, pch = 21, 
                    col = 3, bg = 7, lwd = 2, cex = 2))
  with(HiLo, points(Northing ~ Easting, pch = 25,  
                    col = 4, bg = 3, lwd = 3, cex = 1.4))
  with(LoHi, points(Northing ~ Easting, pch = 24,
                    col = 3, bg = 6, lwd = 3, cex = 1.4))
  mtext(varz[i],3,-2,adj=0.5,font=2,cex=2)
  mtext(paste0("(",letters[i],")"),3,-2,adj=0.98,font=2,cex=2)
}
rect(400090,6467855,400590,6467930,col="#FFFFFF80", border = "#B0B0B080")
legend("bottomright", bty = "n", cex = 1.6, ncol=3,
       legend = c("High-High", "Low-Low","High-Low","Low-High","Samples"),
       pch = c(22, 21, 25,24,3), pt.cex = c(1.8,2,1.4,1.4,0.85), 
       pt.lwd = c(2,2,2,2,1),
       col = c(2,3,4,3,"grey65"), pt.bg = c(6,7,3,6,NA), text.col = 1,
       box.col = "#b0b0b0", box.lwd = 1.3, bg = "#E8E8E8",
       x.intersp = 0.8, inset=c(0.04,0))
text(400340,6467935,pos=3,labels="Spatial autocorrelation key", font=3,
     cex=1.6, offset = 0.15)
par(mfrow=c(1,1), mar=c(3,3,0.5,0.5), oma=c(0,0,0,0), mgp=c(1.6,0.2,0), 
    tcl=0.3,lend="square", ljoin="mitre")

```

```{r remove-temps-2, message=FALSE, warning=FALSE, include=FALSE}
rm(list = c("data_temp","Coords","mI","locMI", "plotdata", "pos0", "neg0",
            "LISA","medREE","HiHi","LoLo","HiLo","LoHi","bw","i","local_moran",
            "var0","varz"))
```

<hr>

## Using the gstat package for geostatistics: variograms, kriging, and visualization

### ├ Make a sf multipoint object from a data frame

In this example (as for Moran's I) we log~10~-transform our variable as its
distribution is highly positively skewed. (Using the untransformed variable
would result in too many apparent upper outliers.)

```{r make-sf-object, paged.print=FALSE, results='hold', echo=FALSE}
data0 <- na.omit(afs1923surf[,c("Easting","Northing","REE")])
colnames(data0) <- c("Easting","Northing","REE"); row.names(data0) <- NULL

# remove duplicate coordinates befoe converting to sf
# (dups <- which(duplicated(paste0(data0[,1],data0[,2]))==TRUE))
# data0 <- data0[-dups,]    ;     rm(dups)
# make sf object
REE_sf <- st_as_sf(data0, coords = c("Easting","Northing"), crs = UTM50S)
# identify rows with duplicate coordinates
duprows <- which(table(unlist(st_equals(REE_sf)))>1)
# add random noise to rows with duplicate coordinates
REE_sf[duprows,] <- st_jitter(REE_sf[duprows,], amount = 2.5)
REE_sf[duprows,]
```

### ├ Plot the spatial object for checking

To quickly check our data, we can simply use the generic `plot()` function
(including the `add=TRUE` option) to look at the spatial location and value of
our variable.

```{r spplot-spdf, fig.height=5.8, fig.width=8.4, fig.align='center', out.width="80%", fig.cap="Quick visualization of sf point data object for log10-transformed REE concentrations in Ashfield Flats sediments 2019-2023.", message=FALSE, warning=FALSE, results='hold', echo=FALSE}
par(oma=c(3,4,1,2), mar=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), 
    lend=2, ljoin=1, tcl=0.3)
plot_tiles(afstiles, axes=T)
with(afr_map, polygon(wetland_E, wetland_N, border = "steelblue", col="#87CEEB80"))
plot(REE_sf, add=T, pch=19, cex=1.5, pal=viridis::plasma(11, alpha=0.7))
```

### ├ Plot a map with range-class symbols

```{r ree-tukey-plot, echo=2:19, fig.width=10, fig.height=7.3, fig.align='center', fig.cap="Map of REE concentrations expressed as symbols for concentration ranges (UQR is 75th percentile, upWsk is upper whisker, med is median, LQR is 25th percentile, loWsk is lower whisker). Data are from Ashfield Flats sediments 2019-2023.", out.width="80%", message=FALSE, warning=FALSE,echo=FALSE}
palette(c("black",plasma(6, alpha=0.65),"grey92", "white"))
par(oma=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), lend=2, ljoin=1, tcl=0.3)
plot_tiles(afstiles)
with(afr_map, polygon(wetland_E, wetland_N, border = "steelblue", col="#87CEEB80"))
axis(1)
mtext(side=1, line=2, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.2)
axis(2, mgp=c(2, 0.5, 0))
mtext(side=2, line=2, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
addnortharrow(pos="topright", border=1, lwd=1, text.col=1,  
              padin=c(0.2,0.2), scale=0.8)
addscalebar(plotepsg = 32750, htin = 0.16, label.cex = 1.2, pos = "bottomright")
# construct and plot range classes in Tukey boxplot thresholds
REE_sf$tukey <- cut(REE_sf$REE, 
    breaks = c(0,boxplot.stats(REE_sf$REE)$stats,999999),
    labels = c("-outliers","loWx-LQR","LQR-med","med-UQR","UQR-upWx","+outliers"))
polygon(af_border$Easting, af_border$Northing, lty=3, lwd=2,
        border="olivedrab")
plot(REE_sf, add=T, pch = 21,
       cex=(c(.4,.52,.68,0.8,1.12,1.44)*2)[REE_sf$tukey], 
       bg=seq(2,7)[REE_sf$tukey], col=seq(2,7)[REE_sf$tukey])
     
legend("topleft", legend = rev(c("-outliers","loWsk to LQR","LQR to med",
                                "med to UQR","UQR to upWsk","+outliers")), 
       cex = 1., pch=21, pt.cex=rev(c(.4,.52,.68,0.8,1.12,1.44))*2, inset = 0.01, 
       bty="o", bg="grey94", y.intersp=1.15, pt.bg=seq(7,2,-1), col=plasma(7), 
       x.intersp = .9)
text(399870, 6468410, 
     labels=expression(bold("REE ranges")), pos = 4, cex = 1)
box()
# par(new=TRUE, fig=c(0,0.07,0.68,0.92), mar=c(0,0,0,0), 
#     col.lab=1, col.axis=1, mgp=c(.9,0.2,0), xaxt="n", yaxt="n", bty = "n", ann = F) 
# boxplot(REE_sf$REE, col = "#A0A0A080");box(which = "plot")
```

Bins for boxplot-categorised-point map in Figure \@ref(fig:ree-tukey-plot):

```{r tukey-map-bins, echo=FALSE, paged.print=FALSE, results='asis'}
tukbins <- data.frame(bins=c("-outliers","loWsk to LQR","LQR to med",
                                "med to UQR","UQR to upWsk","+outliers"),
      logfrom=signif(c(-4,log10(boxplot.stats(REE_sf$REE)$stats)),3),
      logto=signif(c(log10(boxplot.stats(REE_sf$REE)$stats),Inf),3),
      from=round(c(-4,boxplot.stats(REE_sf$REE)$stats),1),
      to=round(c(boxplot.stats(REE_sf$REE)$stats,Inf),1))
knitr::kable(tukbins, table.attr = "class=\"striped\"",format="html",
             caption="Tukey boxplot bins defining ranges for range-class map above.")
```

<p>&nbsp;</p>

----------------------

## Variograms and Kriging

### ├ Make a binned simple variogram object

```{r make-variogram-object, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap="Plot of experimental binned variogram for REE in Ashfield Flats sediments 2019-2023.", fig.height=3.5, fig.width=4.5, out.width="50%"}
par(mar = c(3,3,1,1), mgp = c(1.7,0.3,0), font.lab=2, tcl=0.3)
variog_REE <- variogram(sqrt(REE)~1, REE_sf, cutoff=150) # , cressie = TRUE
attz <- attributes(variog_REE)
plot(variog_REE$dist, variog_REE$gamma, xlab = "Distance (m)", pch=19, 
     col="blue2", ylab = expression(bold(paste("Semivariance (",gamma,")"))),
     ylim = c(0,max(variog_REE$gamma)))
text(par("usr")[2], 0.015, pos=2,
     labels=paste(attz$what, "with", length(attz$boundaries)-1, "bins"))
rm(attz)
```

```{r print-variog-obj, paged.print=FALSE, echo=FALSE}
vgr <- variog_REE; vgr[,2] <- round(vgr[,2],1); vgr[,3] <- round(vgr[,3],3)
vgr[,1:3]; rm(vgr)
```

### ├ Fit a variogram model using weighted least squares

```{r fit-variogram-model, fig.height=3, fig.width=4, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold', echo=FALSE}
vmodel_REE <- fit.variogram(variog_REE, 
                              model = vgm(psill = NA, model = "Gau", 
                                          nugget = 0.1, 
                                          cutoff = max(variog_REE$dist)))
vmodel_REE
```

```{r plot-exp-model-variograms, fig.height=3.6, fig.width=5, fig.align='center', fig.cap="Plot of experimental binned variogram, and exponential variogram model, for REE in Ashfield Flats sediments 2019-2023.", out.width="50%", results='hide', echo=FALSE}
par(mar = c(4,4,1,1), mgp = c(1.7,0.3,0), tcl=0.25, font.lab=2)
with(variog_REE, plot(dist, gamma, ylim = c(0, max(gamma)), cex.lab=1.2, 
                   lwd=2, xlab = "Distance (m)", 
                   ylab = expression(bold(paste("Semivariance (",gamma,")")))))
# plot model
x0 <- variog_REE$dist
y0 <- vmodel_REE$psill[1] + 
  vmodel_REE$psill[2]*(1 - exp(-1*x0/vmodel_REE$range[2]))
lines(x0,y0,col=2)
legend("bottomright", bty = "o", inset=c(0.02,0.04), box.col = "grey", cex=1,
       pch=c(1,NA), pt.lwd=2, lwd=c(NA,1), col=c(1,2), pt.cex=1.1,
       legend = c("Empirical variogram",
                  paste(vmodel_REE$model[2],"variogram model")))
text(150,7,pos=2,offset=0, cex=0.95, font=2, family='mono', col = 2,
     labels=paste(
       " nugget =",signif(vmodel_REE$psill[1],3),"\n",
       "partial sill =", signif(vmodel_REE$psill[2],3),"\n",
       "range =", signif(vmodel_REE$range[2],3),"\n",
       "practical range =",signif(vmodel_REE$range[2]*sqrt(3),3),"" 
  ))
# NOTE factor (range -> practical range) depends on model
# - For an exponential model, practical range = 3 x range
# For a Gaussian model, practical range = sqrt(3) x range
# For a spherical model, practical range ≃ 0.8 x range

```

### ├ Perform kriging interpolation

### ├ first make a grid mask

```{r make-grid-masked-to-AFR-sampling-zones, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
afgrid <- st_make_grid(REE_sf, cellsize = c(5,5), what="corners")
# afgrid <- SpatialPoints(grid0, proj4string = UTM50S)

afpoly <- st_polygon(list(as.matrix(af_border[,c(1,2)])))
afpoly <- st_sfc(afpoly)
st_crs(afpoly) <- UTM50S

af_interp_grid <- st_intersection(afpoly,afgrid)
cat("Prediction grid:\n----------------\n"); af_interp_grid
```

```
    _   _    ____    _______   ______     
   | \ | |  / __ \  |__   __| |  ____|  _ 
   |  \| | | |  | |    | |    | |__    (_)
   | . ` | | |  | |    | |    |  __|      
   | |\  | | |__| |    | |    | |____   _ 
   |_| \_|  \____/     |_|    |______| (_)
                         _      __  _____    ______   ______  __      _ 
                        | |    / / |  __ \  |  ____| |  ____| \ \    | |
    ___    __ _   _ __  | |_  | |  | |__) | | |__    | |__     | |   | |
   / __|  / _` | | '__| | __| | |  |  _  /  |  __|   |  __|    | |   | |
   \__ \ | (_| | | |    | |_  | |  | | \ \  | |____  | |____   | |   |_|
   |___/  \__, | |_|     \__| | |  |_|  \_\ |______| |______|  | |   (_)
             | |               \_\                            /_/       
             |_|                                                        
```

### ├ Krige to grid
```{r krige-to-grid, results='hold', echo=FALSE}
kriged_REE <- krige(formula = sqrt(REE)~1, locations = REE_sf, 
                      newdata = af_interp_grid, model = vmodel_REE)
summary(kriged_REE)
cat("_____________________________________________________________________\n\n")
idwt <- 1.5
idw_REE <- idw(formula = sqrt(REE)~1, locations = REE_sf, 
                      newdata = af_interp_grid, idp = idwt)
idw_REE$var1.var <- NULL
summary(idw_REE)
```

### ├ Simple plot of kriging output

We can quickly use the `plot()` function on our `sf` object to visualise the 
kriging predictions and variance, but without a background map.

```{r spplot-kriging-output, fig.width=8, fig.height=5.85, fig.align='center', echo=2, fig.cap="Plots of simple kriging predictions and variance for log-transformed REE in Ashfield Flats sediments 2019-2023.", out.width="75%", warning=FALSE, error=FALSE, results='hold', echo=FALSE}
plot(kriged_REE, pch=15, nbreaks=32, pal=mako(32))
```

### ├ Plot a map with overlay of the kriging predictions

```{r map-krig-mod-pred, fig.height=16, fig.width=12, fig.align='center', fig.cap="Map showing kriging predictions for REE in Ashfield Flats sediments 2019-2023 by (a) kriging, (b) inverse distance weighted interpolation. Sample points are tiny dots, drains are shown with solid lines, and seasonal wetland pond outlines are semitransparent and dashed.", message=FALSE, warning=FALSE, echo=FALSE}
par(mfrow=c(2,1), oma=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), lend=2, ljoin=1, 
    tcl=0.3, new=FALSE)
nclr <- 52
plot_tiles(afstiles)
with(afr_map, polygon(wetland_E, wetland_N, border = "steelblue", col="#87CEEB80"))
axis(1, mgp=c(2, 0.3, 0), labels=F)
axis(2, mgp=c(2, 0.3, 0), cex.axis = 1.5)
mtext(side=2, line=1.5, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.8)
plot(idw_REE, add=T, pch=15, cex=1.1, breaks=nclr, pal=magma(nclr))
with(afr_map, lines(drain_E, drain_N, col="#00b0E060", lwd=3))
with(afr_map, 
     polygon(wetland_E, wetland_N, border = "#4466E060", col="#00000000", 
             lwd=4, lty="longdash", lend="round", ljoin="round"))
addnortharrow(border=1, lwd=1, text.col=1, padin=c(0.1,0.1), scale = 1.4)
addscalebar(plotepsg = 32750, pos = "bottomleft",htin=0.22, label.cex = 1.8)
points(afs1923surf[,c("Easting","Northing")], pch=19, cex=0.3, col="white")
text(399970,6467930, labels="Derbarl Yerrigan\n(Swan River)", 
     font=3, cex=1.4, col="darkcyan")
box() ; mtext(paste0("(b) IDW (weight ",idwt,")"),3,-2,font=2,cex=1.8,adj=0.025)

plot_tiles(afstiles)
with(afr_map, polygon(wetland_E, wetland_N, border = "steelblue", col="#87CEEB80"))
axis(1, mgp=c(2, 0.3, 0), cex.axis = 1.5)
mtext(side=1, line=1.8, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.8)
axis(2, mgp=c(2, 0.3, 0), cex.axis = 1.5)
mtext(side=2, line=1.5, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.8)
plot(kriged_REE, add=T, pch=15, cex=1.1, breaks=nclr, pal=magma(nclr))
with(afr_map, lines(drain_E, drain_N, col="#00b0E060", lwd=3))
with(afr_map, 
     polygon(wetland_E, wetland_N, border = "#6644E090", col="#00000000", 
             lwd=4, lty="longdash", lend="round", ljoin="round"))
addnortharrow(border=1, lwd=1, text.col=1, padin=c(0.1,0.1), scale = 1.4)
addscalebar(plotepsg = 32750, pos = "bottomleft",htin=0.22, label.cex = 1.8)
points(afs1923surf[,c("Easting","Northing")], pch=19, cex=0.3, col = "white")
text(399970,6467930, labels="Derbarl Yerrigan\n(Swan River)", 
     font=3, cex=1.4, col="darkcyan")
box() ; mtext(paste0("(a) Kriging (",vmodel_REE$model[2]," model)"),3,-2,
              font=2,cex=1.8,adj=0.025)

par(new=TRUE, fig=c(0.5,0.95,0.0,0.05), mar=c(2.5,0.1,0.1,0.1),
    font.lab=2, mgp=c(2,0.4,0), tcl=-0.2)
legrng <- range(st_drop_geometry(kriged_REE[,1]))
plot(legrng,c(0,2),type="n",xaxt="n",yaxt="n",xlab="",ylab="",xaxs="i")
points(seq(legrng[1],legrng[2],l=nclr), 
       rep(1,nclr), col=magma(nclr),pch=15, cex=2.8)
axis(1, at=sqrt(c(0,20,40,100,200,400,1000)), labels=c(0,20,40,100,200,400,1000),
     mgp=c(1.5,0.4,0)) ; box()
mtext(expression(bold(paste(Sigma,"REE"))),1,1.3,cex=1.5)
```

