---
title: "Ashfield 2019-2022 spatial analyses"
output: 
  bookdown::html_document2:
    fig_width: 6.4
    fig_height: 4.8
    number_sections: no
    self_contained: no
    df_print: default
    fig_caption: yes
    smart: no
---

```{r load packages and environment but do not show, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}
library(sp)
library(lctools)
library(gstat)
library(sf)
library(maptiles)
library(prettymapr)
library(fields)
library(viridis)
library(ncf)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
git <- "https://github.com/Ratey-AtUWA/Ashfield-REE/raw/main/"
```

```{r importing the data, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
afs1922 <- read.csv(paste0(git,"afs1922.csv"), stringsAsFactors = TRUE)
afs1922$Year <- as.factor(afs1922$Year)
af_border <- read.csv(paste0(git,"afr_hull.csv"))
afs1922surf <- droplevels(afs1922[-which(afs1922$Depth_upper_cm > 1 & 
                                           afs1922$Type=="Soil"),])
```

# Making our base map of Ashfield Flats

```{r define CRS, warning=FALSE, error=FALSE, results='hold', echo=FALSE}
LongLat <- CRS("+proj=longlat +ellps=WGS84 
           +datum=WGS84 +no_defs") # uses Earth ellipsis specs from WGS84 datum
UTM50 <- CRS("+proj=utm +zone=50 +south") # just for Zone 50, S hemisphere!
```

### ├ Map extent object

```{r make map extent object, warning=FALSE, error=FALSE, results='hold', echo=FALSE}
extent <- st_as_sf(x = data.frame(x = c(399860,400580), 
                                  y = c(6467870,6468350)), 
                   coords = c("x", "y"), crs = UTM50)
st_coordinates(extent)
```

### ├ Getting and plotting the map tile data

```{r basemap, fig.height=5.85, fig.width=8, fig.align='center', fig.cap="Map of Ashfield Flats Reserve and adjacent area (UTM projection, Zone 50S, EPSG:32750) used subsequently as the base map for spatial analyses. Generated using the maptiles R package, with OpenStreetMap tiles.", out.width="75%", warning=FALSE, message=FALSE, results='hide', echo=FALSE}
aftiles <- get_tiles(extent, provider = "OpenStreetMap.HOT", 
                     crop = TRUE, zoom = 17) # make map object

par(oma=c(3,3,1,1), mar=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), 
    lend=2, ljoin=1, tcl=0.3)
plot(aftiles)
axis(1)
mtext(side=1, line=2, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.2)
axis(2)
mtext(side=2, line=2, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
with(afs1922surf, points(Easting, Northing, pch = 3, cex = 0.75, col = "navy"))
with(af_border, lines(Easting,Northing, lty = 3, col = "darkgreen"))
addnortharrow()
addscalebar(plotepsg = 32750, pos = "topleft")
box()
```

#### **Note**: in this document REE ($\equiv \sum$REE) = Ce + La + Nd + Gd.
$\sum$REE does not include yttrium (Y) at present, but it could using \
`afs1922$YREE <- afs1922$REE + afs1922$Y`.

## Spatial Autocorrelation

### ├ Calculate Global Moran's I

```{r global Moran, echo=TRUE, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
var0 <- "REE" # choose the variable of interest

# Calculate global Moran's I
data_temp <- na.omit(afs1922surf[,c("Easting", "Northing", var0)])
Coords <- cbind(data_temp$Easting, data_temp$Northing)
bw <- 8 # bw = bandwidth, = number of nearest point for calculation
mI <- moransI(Coords, bw, data_temp[,3])
cat("Global Moran's I for ",var0,"; from the NGSA (af only), topsoil, -2mm fraction\n", sep="");
t(as.matrix(mI[c(2,4:7)]))
rm(list = c("data_temp","Coords","bw","mI","var0")) # remove temporary objects
```

### ├ Plot local Moran's I

```{r local-Moran, message=FALSE, warning=FALSE, fig.height=5.85, fig.width=8, fig.align='center', out.width="80%", fig.cap="Map of Local Moran's I for REE concentrations in Ashfield Flats sediments 2019-2022. The Global Moran's I parameter is also shown beneath the legend.", results='hide', echo=FALSE}
var0 <- "REE" # choose the variable of interest
data_temp <- na.omit(afs1922surf[,c("Easting", "Northing", var0)])
Coords <- cbind(data_temp$Easting, data_temp$Northing)
mI <- moransI(Coords, 8, log10(data_temp[,3])) # log10 minimises skewness
local_moran <- l.moransI(Coords, 8, log10(data_temp[,3]), scatter.plot = FALSE)
plotdata <- data.frame(Easting=Coords[,1], Northing=Coords[,2], 
                       MoranI=local_moran$Ii, p_value=local_moran$p.value)
pos0 <- subset(plotdata, plotdata$MoranI>0 & plotdata$p_value<=0.05)
neg0 <- subset(plotdata, plotdata$MoranI<0 & plotdata$p_value<=0.05)
#
par(oma=c(3,3,1,1), mar=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), 
    lend=2, ljoin=1, tcl=0.3, tcl = 0.5, lwd = 1)
plot(aftiles)
axis(1)
mtext(side=1, line=2, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.2)
axis(2)
mtext(side=2, line=2, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
addnortharrow(pos="topright", border=1, lwd=1, text.col=1,  
              padin=c(0.1,0.2), scale=1.2)
addscalebar(plotepsg = 32750)
box()
points(Coords, pch=3, cex=0.5, col = "grey67")
with(pos0, symbols(Easting, Northing, circles = 15*sqrt(MoranI*0.02), 
                   inches = F, fg = "blue3", bg = "#00008080", add= TRUE))
with(neg0, symbols(Easting, Northing, squares = 15*sqrt(MoranI*-0.04), 
                   inches = F, fg = "red3", bg = "#80000080", add= TRUE))
# rect(107,-18,120,-12.5, col="#E4DDB7", border = "white", lwd = 2)
legend("bottomright", bty = "o", cex = 1, title.cex = 0.9,
       legend = c("Positive I", "Negative I",
                  "", NA),
       title = paste0("Local Moran's I for log(",var0, 
                      ") in\nafs1922 (p \u2264 0.05 only)"),
       pch = c(21, 22,NA,NA), pt.cex = c(2, 1.7,NA,NA),
       col = c("blue3", "red3"), pt.bg = c("#00008080", "#80000080"),
       title.col = "navy", y.intersp = 0.9)
text(400600,6467860, 
     labels = paste("Global Moran's I =", signif(as.numeric(mI[2]),3),
                               "\nRandomization p =", signif(as.numeric(mI[7]),3)), 
     pos = 2)
```

### ├ Plot 'LISA'

```{r lisa-plot, message=FALSE, warning=FALSE, fig.height=5.85, fig.width=8, fig.align='center', out.width="80%", fig.cap="Map of LISA for REE concentrations in the Ashfield Flats sediment data, 2019-2022.", results='hide',echo=FALSE}
data_temp <- na.omit(as.data.frame(cbind(afs1922surf$Easting,afs1922surf$Northing,
                                         afs1922surf$REE)))
colnames(data_temp) <- c("Easting","Northing","REE")


Coords <- cbind(data_temp$Easting, data_temp$Northing)
bw <- 8
locMI <- l.moransI(Coords,bw,data_temp$REE, scatter.plot = F)
LISA <- as.data.frame(cbind(Coords[,1:2], log10(data_temp$REE), 
                            locMI$Ii, 
                            locMI$p.value))
colnames(LISA) <- c("Easting", "Northing", "REE", "MoranI", "p_value") # ;head(LISA)
medREE <- median(LISA$REE, na.rm = TRUE)
HiHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA$REE>=medREE)
LoLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA$REE<medREE)
HiLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA$REE>=medREE)
LoHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA$REE<medREE)
#
par(oma=c(3,3,1,1), mar=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), 
    lend=2, ljoin=1, tcl=0.3, tcl = 0.5, lwd = 1)
palette(c("black","#400000","blue3","orange","skyblue",
          "#FF8080","#8080FF","#FFA50040","#87CEEB40","white"))
plot(aftiles)
axis(1)
mtext(side=1, line=1.6, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.2)
axis(2)
mtext(side=2, line=1.6, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
addnortharrow(pos="topright", border=1, lwd=1, text.col=1,  
              padin=c(0.1,0.2), scale=1.2)
addscalebar(plotepsg = 32750)
box()
with(data_temp, points(Easting, Northing, pch=3, cex = 0.7, col="grey"))
with(HiHi, points(Northing ~ Easting, pch = 22,
                  col = 2, bg = 6, lwd = 2, cex = 1.8))
with(LoLo, points(Northing ~ Easting, pch = 21, 
                  col = 3, bg = 7, lwd = 2, cex = 2))
with(HiLo, points(Northing ~ Easting, pch = 25,  
                  col = 2, bg = 7, lwd = 3, cex = 1.4))
with(LoHi, points(Northing ~ Easting, pch = 24,
                  col = 3, bg = 6, lwd = 3, cex = 1.4))
legend("bottomright", bty = "o", inset = 0.015, cex = 1, ncol = 3,
       legend = c("High-High", "Low-Low","High-Low","Low-High","Sample points"),
       pch = c(22, 21, 25,24,3), pt.cex = c(1.8,2,1.4,1.4,0.7), 
       pt.lwd = c(2,2,3,3,1),
       col = c(2,3,2,3,"grey"), pt.bg = c(6,7,7,6,NA), text.col = 1,
       box.col = "#b0b0b0", box.lwd = 1.3, bg = "#E8E8E8", 
       title="\nSpatial autocorrelation classes for REE",
       y.intersp = 1.1)
```

### ├ LISA maps for multiple variables

```{r multi-LISA, fig.height=10.6, fig.width=7.8, fig.align='center', fig.cap="LISA autocorrelation maps for REE, Al, Fe, S, As, Cu, Pb, and Zn concentrations in the Ashfield Flats sediment data, 2019-2022.", message=FALSE, warning=FALSE, echo=FALSE}
par(mfrow=c(4,2), mar=c(0,0,0,0), oma=c(3,3,0.5,0.5), mgp=c(1.6,0.2,0), 
    tcl=0.3,lend="square", ljoin="mitre")
palette(c("black","#400000","blue3","orange","skyblue",
          "#FF8080","#8080FF","#FFA50040","#87CEEB40","white"))
varz <- c("REE","Al","Fe","S","As","Cu","Pb","Zn")
for (i in 1:length(varz)){
  data_temp <- na.omit(as.data.frame(cbind(afs1922$Easting,afs1922$Northing,
                                         afs1922[,varz[i]])))
  colnames(data_temp) <- c("Easting","Northing",varz[i])

  Coords <- cbind(data_temp$Easting, data_temp$Northing)
  bw <- 8
  locMI <- l.moransI(Coords,bw,data_temp[,varz[i]], scatter.plot = F)
  LISA <- as.data.frame(cbind(Coords[,1:2], log10(data_temp[,varz[i]]), 
                            locMI$Ii, 
                            locMI$p.value))
  colnames(LISA) <- c("Easting", "Northing", var0, "MoranI", "p_value")
  medREE <- median(LISA[,var0], na.rm = TRUE)
  HiHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA[,var0]>=medREE)
  LoLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI>0&LISA[,var0]<medREE)
  HiLo <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA[,var0]>=medREE)
  LoHi <- subset(LISA, LISA$p_value<=0.05&LISA$MoranI<0&LISA[,var0]<medREE)

  plot(aftiles)
  if(i > 6.5) {
    axis(1)
    mtext(side=1, line=1.6, text="Easting (UTM Zone 50, m)", font=2, cex=1)
    }
  if(round(i/2,0) != (i/2)) {
    axis(2)
    mtext(side=2, line=1.6, text="Northing (UTM Zone 50, m)", font=2, cex=1)
  }
  axis(1, labels=F); axis(2, labels = F)
  addnortharrow(pos="topleft", border=1, lwd=1, text.col=1,  
              padin=c(0.05,0.05), scale = 0.8)
  addscalebar(plotepsg = 32750, label.cex = 1.2)
  box()
  with(data_temp, points(Easting, Northing, pch=3, cex = 0.7, col="grey"))
  with(HiHi, points(Northing ~ Easting, pch = 22,
                  col = 2, bg = 6, lwd = 2, cex = 1.8))
  with(LoLo, points(Northing ~ Easting, pch = 21, 
                  col = 3, bg = 7, lwd = 2, cex = 2))
  with(HiLo, points(Northing ~ Easting, pch = 25,  
                  col = 2, bg = 7, lwd = 3, cex = 1.4))
  with(LoHi, points(Northing ~ Easting, pch = 24,
                  col = 3, bg = 6, lwd = 3, cex = 1.4))
  mtext(varz[i],3,-1.5,adj=0.98,font=2)
}
legend("bottomright", bty = "n", cex = 1.2, ncol=3,
       legend = c("High-High", "Low-Low","High-Low","Low-High","Samples"),
       pch = c(22, 21, 25,24,3), pt.cex = c(1.8,2,1.4,1.4,0.7), 
       pt.lwd = c(2,2,2,2,1),
       col = c(2,3,2,3,"grey"), pt.bg = c(6,7,7,6,NA), text.col = 1,
       box.col = "#b0b0b0", box.lwd = 1.3, bg = "#E8E8E8", 
       title=expression(italic("Spatial autocorrelation key")),
       x.intersp = 0.8)
par(mfrow=c(1,1), mar=c(3,3,0.5,0.5), oma=c(0,0,0,0), mgp=c(1.6,0.2,0), 
    tcl=0.3,lend="square", ljoin="mitre")

```

```{r remove temps 2, message=FALSE, warning=FALSE, include=FALSE}
rm(list = c("data_temp","Coords","mI","locMI", "plotdata", "pos0", "neg0",
            "LISA","medREE","HiHi","LoLo","HiLo","LoHi","bw","i","local_moran",
            "var0","varz"))
```

<hr>

## Using the gstat package for geostatistics: variograms, kriging, and visualization

### ├ Make a 'SpatialPointsDataFrame' object from a data frame

In this example (as for Moran's I) we $log_{10}$-transform our variable as its
distribution is highly positively skewed. (Using the untransformed variable
would result in too many apparent upper outliers.)

```{r make SpatialPointsDataFrame object, paged.print=FALSE, results='hold', echo=FALSE}
var0 <- "REE" # choose variable
data0 <- na.omit(afs1922surf[,c("Easting","Northing",var0)])
colnames(data0) <- c("Easting","Northing",var0)
spname <- paste0(var0,"_sp")
var1 <- as.data.frame(log10(data0[,var0])); colnames(var1) <- var0
assign(spname, 
       SpatialPointsDataFrame(coords = data0[,c("Easting","Northing")],
                              data = var1,
                              proj4string = UTM50))
REE_sp <- REE_sp[-zerodist(REE_sp)[,1],]
summary(REE_sp)
```

### ├ Plot the spatial object for checking

To quickly check our data, we use the function bubble()} from the
sp} package to make a *bubble map* of our variable, where the symbol
*area* is proportional to the variable value (in this case, $log_{10}$ gadolinium
concentration).

```{r spplot-spdf, fig.height=5, fig.width=8, fig.align='center', out.width="80%", fig.cap="Visualization of spatial point data object for log10-transformed REE concentrations in Ashfield Flats sediments 2019-2022.", message=FALSE, warning=FALSE, results='hold', echo=FALSE}
bubble(REE_sp, var0, col=c("#ff000088", "#0000ff88"), main = "", 
       scales = list(draw = TRUE), xlab = "Easting", ylab = "Northing")
```

### ├ Plot a map with range-class symbols

```{r ree-tukey-plot, echo=2:19, fig.width=8, fig.height=5.85, fig.align='center', fig.cap="Map of REE concentrations expressed as symbols for concentration ranges (UQR is 75th percentile, upWsk is upper whisker, med is median, LQR is 25th percentile, loWsk is lower whisker). Data are from Ashfield Flats sediments 2019-2022.", out.width="80%", message=FALSE, warning=FALSE,echo=FALSE}
palette(c("black",rev(c("#EE0000C0", "#EE7621C0", "#EEC900C0", "#40E0D0C0", 
                        "#4870F0C0", "#9932CCC0")),"grey92", "white"))
par(oma=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), lend=2, ljoin=1, tcl=0.3)
plot(aftiles)
axis(1)
mtext(side=1, line=2, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.2)
axis(2, mgp=c(2, 0.5, 0))
mtext(side=2, line=2, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
addnortharrow(pos="topright", border=1, lwd=1, text.col=1,  
              padin=c(0.2,0.2), scale=0.8)
addscalebar(plotepsg = 32750, htin = 0.16, label.cex = 1.2, pos = "bottomright")
# construct and plot range classes in Tukey boxplot thresholds
REE_sp@data$tukey <- cut(REE_sp@data$REE, 
    breaks = c(0,boxplot.stats(REE_sp@data[,1])$stats,999999),
    labels = c("-outliers","loWx-LQR","LQR-med","med-UQR","UQR-upWx","+outliers"))
polygon(af_border$Easting, af_border$Northing, lty=3, lwd=2,
        border="olivedrab")
points(REE_sp@coords, pch = 21,
       cex=(c(.5,.65,.85,1,1.4,1.8)*2)[REE_sp@data$tukey], 
       bg=seq(2,7)[REE_sp@data$tukey])
     
legend(399870, 6468370, legend = rev(c("-outliers","loWsk to LQR","LQR to med",
                                "med to UQR","UQR to upWsk","+outliers")), 
       cex = 1., pch=21, pt.cex=rev(c(.5,.65,.85,1,1.4,1.8))*2, inset = 0.07, 
       bty="n", y.intersp=1.15, pt.bg=seq(7,2,-1), x.intersp = .9)
text(399870, 6468410, 
     labels=expression(bold("REE ranges")), pos = 4, cex = 1)
box()
par(new=TRUE, fig=c(0,0.07,0.68,0.92), mar=c(0,0,0,0), 
    col.lab=1, col.axis=1, mgp=c(.9,0.2,0), xaxt="n", yaxt="n", bty = "n", ann = F) 
boxplot(REE_sp@data$REE, col = "#A0A0A080");box(which = "plot")
```

Bins for Figure \@ref(fig:ree-tukey-plot):

## Variograms and Kriging
```{r tukey-map-bins, echo=FALSE}
(data.frame(bins=c("-outliers","loWsk to LQR","LQR to med",
                                "med to UQR","UQR to upWsk","+outliers"),
      logfrom=signif(c(-4,boxplot.stats(REE_sp@data[,1])$stats),3),
      logto=signif(c(boxplot.stats(REE_sp@data[,1])$stats,Inf),3),
      from=round(10^c(-4,boxplot.stats(REE_sp@data[,1])$stats),1),
      to=round(10^c(boxplot.stats(REE_sp@data[,1])$stats,Inf),1)))
```

### ├ Make a binned simple variogram object

```{r make variogram object, message=FALSE, warning=FALSE, echo=FALSE, fig.align='center', fig.cap="Plot of experimental binned variogram for REE in Ashfield Flats sediments 2019-2022.", fig.height=3.5, fig.width=4.5, out.width="50%"}
par(mar = c(3,3,1,1), mgp = c(1.7,0.3,0), font.lab=2, tcl=0.3)
variog_REE <- variogram(REE~1, REE_sp) # , cressie = TRUE
attz <- attributes(variog_REE)
plot(variog_REE$dist, variog_REE$gamma, xlab = "Distance (m)", pch=19, 
     col="blue2", ylab = expression(bold(paste("Semivariance (",gamma,")"))),
     ylim = c(0,max(variog_REE$gamma)))
text(250, 0.015, pos=2,
     labels=paste(attz$what,"with",length(attz$boundaries)-1,"bins"))
rm(attz)
```

```{r print-variog-obj, paged.print=FALSE, echo=FALSE}
vgr <- variog_REE; vgr[,2] <- round(vgr[,2],1); vgr[,3] <- round(vgr[,3],3)
vgr[,1:3]; rm(vgr)
```

### ├ Fit a variogram model using weighted least squares

```{r fit variogram model, fig.height=3, fig.width=4, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold', echo=FALSE}
vmodel_REE <- fit.variogram(variog_REE, 
                              model = vgm(psill = NA, model = "Exp", 
                                          nugget = 0.1, 
                                          cutoff = max(variog_REE$dist)))
vmodel_REE
```

```{r plot-exp-model-variograms, fig.height=3.6, fig.width=5, fig.align='center', fig.cap="Plot of experimental binned variogram, and exponential variogram model, for REE in Ashfield Flats sediments 2019-2022.", out.width="50%", results='hide', echo=FALSE}
par(mar = c(4,4,1,1), mgp = c(1.7,0.3,0), tcl=0.25, font.lab=2)
with(variog_REE, plot(dist, gamma, ylim = c(0, max(gamma)), cex.lab=1.2, 
                   lwd=2, xlab = "Distance (m)", 
                   ylab = expression(bold(paste("Semivariance (",gamma,")")))))
# plot model
x0 <- variog_REE$dist
y0 <- vmodel_REE$psill[1] + 
  vmodel_REE$psill[2]*(1 - exp(-1*x0/vmodel_REE$range[2]))
lines(x0,y0,col=2)
legend("bottomright", bty = "n", cex=1, pch=c(1,NA), pt.lwd=2, lwd=c(NA,1), 
       col=c(1,2), pt.cex=1.1,legend = c("Empirical variogram",
                  paste(vmodel_REE$model[2],"variogram model")))
text(260,0.05,pos=2,offset=0, cex=0.95, family='mono', col = 2,
     labels=paste(
       " nugget =",signif(vmodel_REE$psill[1],3),"\n",
       "partial sill =", signif(vmodel_REE$psill[2],3),"\n",
       "range =", signif(vmodel_REE$range[2],3),"\n",
       "practical range =",signif(vmodel_REE$range[2]*3,3),"" 
  ))
# NOTE factor (range -> practical range) depends on model
```

### ├ Perform kriging interpolation

### ├ first make a grid mask

```{r make grid masked to AFR sampling zones, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
grid0 <- expand.grid(seq(round(min(REE_sp@coords[,1]),1), 
                            round(max(REE_sp@coords[,1]),1), 5),
                        seq(round(min(REE_sp@coords[,2]),1), 
                            round(max(REE_sp@coords[,2]),1), 5))
afgrid <- SpatialPoints(grid0, proj4string = UTM50)

irregpoly <- Polygon(af_border[,c(1,2)], hole=F)
irregPolys = Polygons(list(irregpoly),1)
gridMask = SpatialPolygons(list(irregPolys), 
                           proj4string = UTM50)
inOrOut <- as.vector(over(afgrid, gridMask))
afgrid <- afgrid[which(inOrOut>0)]
cat("Prediction grid:\n"); summary(afgrid)
```

### ├ Krige to grid
```{r krige-to-grid, results='hold', echo=FALSE}
kriged_REE <- krige(formula = REE~1, locations = REE_sp, 
                      newdata = afgrid, model = vmodel_REE)
summary(kriged_REE)
cat("_____________________________________________________________________\n\n")
idw_REE <- idw(formula = REE~1, locations = REE_sp, 
                      newdata = afgrid, idp = 2)
idw_REE@data$var1.var <- NULL
summary(idw_REE)
```

### ├ Simple plot of kriging output

We can then us the spplot() function from sp to visualise the 
kriging predictions and variance, but without a background map.

```{r spplot-kriging-output, fig.width=8, fig.height=5.85, fig.align='center', echo=2, fig.cap="Plots of simple kriging predictions and variance for log-transformed REE in Ashfield Flats sediments 2019-2022.", out.width="75%", warning=FALSE, error=FALSE, results='hold', echo=FALSE}
par(oma=c(0,0,0,0))
spplot(kriged_REE, colorkey=T, pch=15, # with kriging variances
    names.attr = c("log10[REE] kriging predictions",
                   "log10[REE] kriging variance"), 
    scales = list(draw = TRUE), col.regions = viridis::mako(100),
    xlab="Longitude", ylab="Latitude")
```

### ├ Plot a map with overlay of the kriging predictions

```{r map-krig-mod-pred, fig.height=10.6, fig.width=8, fig.align='center', fig.cap="Map showing kriging predictions for REE in Ashfield Flats sediments 2019-2022. Sample points are tiny dots.", message=FALSE, warning=FALSE, echo=FALSE}
par(mfrow=c(2,1), oma=c(4,4,1.5,1.5), mgp=c(1.4,0.3,0), lend=2, ljoin=1, tcl=0.3)
plot(aftiles)
axis(1, mgp=c(2, 0.3, 0), labels=F)
# mtext(side=1, line=1.5, text="Easting (UTM Zone 50, m)",
#       font=2, cex=1.2)
axis(2, mgp=c(2, 0.3, 0), cex.axis = 1)
mtext(side=2, line=1.5, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
axtx <- c(1,2,3,5,10,20,30,50,100,200,300,500,1000)
colgrad <- gsub("FF","B0", viridis::mako(128)[round(seq(1, 128, l=6),0)])
quilt.plot(afgrid@coords[,1], afgrid@coords[,2],kriged_REE@data[,1],
           add = T, horizontal=T, 
           col = colorRampPalette(colgrad, alpha = TRUE)(64),
           legend.lab = expression(bold("Kriging prediction of REE (mg/kg)")), 
           legend.mar = 6, legend.cex = 1, legend.line = 1.2,
           axis.args = list(at=log10(axtx), labels=axtx, mgp=c(1.2,0.1,0)))
addnortharrow(border=1, lwd=1, text.col=1, padin=c(0.1,0.1), scale = 0.9)
addscalebar(plotepsg = 32750, pos = "topleft")
points(afs1922surf[,c("Easting","Northing")], pch=".", col = "yellow")
box() ; mtext("(a)",3,-1.5,font=2,cex=1.5)

plot(aftiles)
axis(1, mgp=c(2, 0.3, 0), cex.axis = 1)
mtext(side=1, line=1.5, text="Easting (UTM Zone 50, m)",
      font=2, cex=1.2)
axis(2, mgp=c(2, 0.3, 0), cex.axis = 1)
mtext(side=2, line=1.5, text="Northing (UTM Zone 50, m)",
      font=2, cex=1.2)
axtx <- seq(0,0.14,0.02)
colgrad <- gsub("FF","80", viridis::cividis(128)[round(seq(1, 128, l=6),0)])
quilt.plot(afgrid@coords[,1], afgrid@coords[,2],kriged_REE@data[,2],
           add = T, horizontal=T, 
           col = colorRampPalette(colgrad, alpha = TRUE)(64),
           legend.lab = expression(bold("Kriging variance in REE (mg/kg)")), 
           legend.mar = 6, legend.cex = 1, legend.line = 1.2,
           axis.args = list(at=axtx, labels=axtx, mgp=c(1.2,0.1,0)))
addnortharrow(border=1, lwd=1, text.col=1, padin=c(0.1,0.1), scale = 0.9)
addscalebar(plotepsg = 32750, pos = "topleft")
points(afs1922surf[,c("Easting","Northing")], pch=".", col="red3")
box() ; mtext("(b)",3,-1.5,font=2,cex=1.5)
```

