---
title: "EDA Ashfield Sediment Data 2019-2022"
output: 
  bookdown::html_document2:
    fig_width: 6.4
    fig_height: 4.8
    number_sections: no
    self_contained: no
    df_print: default
    fig_caption: yes
    smart: no
---

This document describes exploratory and some detailed data analysis

```{r setup packages, warning = FALSE, message = FALSE, results='hide', echo=FALSE}
load(".RData")
library(prettymapr)
library(ggmap)
library(sf)
library(sp)
library(maptiles)
library(TeachingDemos)
library(geosphere)
library(RcmdrMisc)
library(flextable)
  set_flextable_defaults(font.family = 'sans', theme_fun = "theme_zebra", 
        font.size = 11, text.align = "right",padding.left = 1, padding.right = 1,
        padding.top = 2, padding.bottom = 2)
library(car)
library(diptest)
library(PMCMRplus)
library(rcompanion)
library(multcompView)
library(factoextra)
library(ggpubr)
library(viridis)
```

```{r define CRS, warning=FALSE, error=FALSE, results='hold', echo=FALSE}
LongLat <- CRS("+proj=longlat +ellps=WGS84 
           +datum=WGS84 +no_defs") # uses Earth ellipsis specs from WGS84 datum
UTM50S <- CRS("+proj=utm +zone=50 +south") # just for Zone 50, S hemisphere!
```

```{r setup data, echo=FALSE}
git <- "https://github.com/Ratey-AtUWA/Ashfield-REE/raw/main/"
afs1922 <- read.csv(paste0(git,"afs1922.csv"), stringsAsFactors = TRUE)
afs1922$Year <- as.factor(afs1922$Year)
afs1922surf <- afs1922[-which(afs1922$Depth_upper_cm > 1 & 
                                           afs1922$Type=="Soil"),]
afs1922surf[which(afs1922surf$Type=="Soil"),"Type"] <- 
  rep("Saltmarsh",length(which(afs1922surf$Type=="Soil")))
afs1922surf <- droplevels(afs1922surf)
afs1922surf$Type <- factor(afs1922surf$Type, 
                 levels = c("Drain_Sed", "Lake_Sed", "Saltmarsh", "Flooded"))
afs1922utm <- st_as_sf(afs1922surf, coords = c("Easting","Northing"), 
                       crs = st_crs(32750))
afs1922ll <- st_transform(afs1922utm, crs = st_crs(4326))
```

```{r annotation data and palettes, echo=FALSE}
afr_map <- read.csv(file=paste0(git,"afr_map_v3.csv"), stringsAsFactors = TRUE)
afr_zones <- read.csv(file=paste0(git,"afr_zones.csv"), stringsAsFactors = TRUE)
pal4lite <- c("black", "purple2", "blue2", "cyan4", "forestgreen", 
              "darkorange", "gold3", "red3", "gray80", "white")
pal4liteTransp <- c("#000000","#912CEEb0","#0000EEb0","#008B8Bb0","#228B22b0",
                    "#CDAD00b0", "#FF8C00b0", "#CD0000b0", "#CCCCCC", "#FFFFFF")
pal4dark <- c("white", "pink1", "orange", "khaki1", "lightgreen", "cadetblue1", 
              "deepskyblue", "plum", "gray80", "black")
```

# REE normalization & pseudo-fractionation

## ├ make PAAS reference data

```{r make-reference-data, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}
PAAS <- data.frame(Element=c("Y", "La","Ce","Pr","Nd","Sm","Eu","Gd",
                             "Tb","Dy","Ho","Er","Tm","Yb","Lu","REE"),
            PAAS=c(27,38,80,8.9,32,5.6,1.1,4.7,0.77,4.4,1,2.9,0.4,2.8,0.43,154.7))
row.names(PAAS) <- PAAS$Element
cat("Crustal average REE concentration in post-Archean Austrlian Shale (PAAS) proxy\n(Taylor & McLennan 1985). *NOTE: in these data, REE = La + Ce + Nd + Gd. \n------------------\n")
print(PAAS)
```

## ├ normalize YREE to PAAS and plot by element
```{r make-PAAS-normalised-data-and-plot, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}
afs1922_PAAS <- afs1922utm[,c(1,6:11,35,27,33,41,52,54)]
afs1922_PAAS$Type <- factor(afs1922_PAAS$Type, 
        levels=c("Drain_Sed","Lake_Sed","Saltmarsh","Flooded"))
varz <- c("La","Ce","Nd","Gd","Y","REE")
for (i in 1:length(varz)){
  afs1922_PAAS[,varz[i]] <- afs1922_PAAS[,varz[i]]/PAAS[varz[i],"PAAS"]
}
plotdata <- 
  stack(st_drop_geometry(afs1922_PAAS[,c("La","Ce","Nd","Gd","Y","REE")]))
plotdata$ind <- factor(plotdata$ind, levels=c("La","Ce","Nd","Gd","Y","REE"))
par(mfrow=c(1,1), mar = c(3,4,1,1), mgp=c(1.7,0.3,0), tcl=0.25, font.lab=2,
    lend=2, ljoin=1)
meanz <- with(plotdata, tapply(values, ind, FUN=mean, na.rm=TRUE))
with(plotdata, boxplot(values ~ ind, xaxt="n",  
                       xlab="", ylab="Normalized REE = sample/PAAS",
                       col = viridis::mako(9, dir=1)[4:9]))
points(c(1:6), as.numeric(meanz), pch=3, col="white", lwd=4)
points(c(1:6), as.numeric(meanz), pch=3, lwd=2)
axis(1, at=1:6, labels=c("La","Ce","Nd","Gd","Y","\u2211REE"), font=2)
abline(h=0.99, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
```

```{r, results='hold'}
ree0 <- with(plotdata, signif(tapply(values, ind, mean, na.rm=T),3))
ree0 <- rbind(ree0,with(plotdata, signif(tapply(values, ind, sd, na.rm=T),3)))
ree0 <- rbind(ree0,with(plotdata, signif(tapply(values, ind, min, na.rm=T),3)))
ree0 <- rbind(ree0,with(plotdata, signif(tapply(values, ind, max, na.rm=T),3)))
row.names(ree0) <- c("Mean","SD","Min","Max")
cat("Statistics for PAAS-normalized REE across all samples & years:\n")
print(ree0)
```

## ├ over all data, are YREE fractionated from each other?
```{r}
with(plotdata, bartlett.test(values~ind))
with(plotdata, oneway.test(values~ind))
with(plotdata, pairwise.t.test(values, ind))
```

Based on these comparisons it seems there is some fractionation. Gd is enriched
which could be due to (1) the 'MREE bulge' commonly observed in reduced
environments, or (2) anthropogenic addition of Gd.

## ├ are YREE significantly differnt from PAAS?

```{r}
with(afs1922_PAAS, t.test(La, mu=1))
with(afs1922_PAAS, t.test(Ce, mu=1))
with(afs1922_PAAS, t.test(Nd, mu=1))
with(afs1922_PAAS, t.test(Gd, mu=1))
with(afs1922_PAAS, t.test(Y, mu=1))
with(afs1922_PAAS, t.test(REE, mu=1))
```

Based on these analyses all YREE differ significantly from PAAS - La, Ce, Nd, Gd greater and Y less. ∑REE is also significantly greater than PAAS.

## ├ Comparisons across environment Types

### ├ ∑REE

```{r REE-normalised-PAAS-by-Type}
s0=1;par(mfrow = c(1,1), mar = c(3.5,3.5,0.5,0.5)*s0, mgp = c(1.6,0.3,0)*s0,
    cex.lab = 1.5*s0, cex.axis=1.4*s0, font.lab = 2, tcl = 0.2*s0, las=1)
with(afs1922_PAAS, boxplot(REE ~ Type, xlab="", xaxt="n", 
     ylab=expression(bold(paste(Sigma,REE[sample]," / ",Sigma,"REE"[PAAS]))), 
     col=c("royalblue","wheat","honeydew3","lightblue"), ylim=c(0,3.8)))
axis(1, at=1:4, labels=c("Drain\nsediment","Lake\nsediment",
             "Saltmarsh\n","Flooded\nwoodland"), mgp=c(3,1.8,0), font=2)
abline(h=0.995, lty="23",col="coral4");abline(h=1.005, lty="23", col="gold")
```

### ├ La-Ce-Nd-Gd

```{r La-Ce-Nd-Gd-norm-PAAS-by-Type, fig.height=10, fig.width=10}
s0=1;par(mfrow = c(3,2), mar = c(3.5,4,0.5,0.5)*s0, mgp = c(1.6,0.3,0)*s0,
    cex.lab = 1.8*s0, cex.axis=1.6*s0, font.lab = 2, tcl = 0.2*s0, las=1)
with(afs1922_PAAS, boxplot(La ~ Type, xlab="", xaxt="n", cex.lab=1.8, 
     ylab=expression(bold(paste(La[sample],"/La"[PAAS]))), ylim=c(0,3.8), 
     col=c("royalblue","wheat","honeydew3","lightblue"), cex=1.5))
axis(1, at=1:4, labels=c("Drain\nsediment","Lake\nsediment","Saltmarsh\n",
        "Flooded\nwoodland"), mgp=c(3,2,0), gap.axis=0)
abline(h=0.98, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
with(afs1922_PAAS, boxplot(Ce ~ Type, xlab="", xaxt="n", cex.lab=1.8, 
     ylab=expression(bold(paste(Ce[sample],"/Ce"[PAAS]))), ylim=c(0,3.8), 
     col=c("royalblue","wheat","honeydew3","lightblue"), cex=1.5))
axis(1, at=1:4, labels=c("Drain\nsediment","Lake\nsediment","Saltmarsh\n",
        "Flooded\nwoodland"), mgp=c(3,2,0), gap.axis=0)
abline(h=0.98, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
with(afs1922_PAAS, boxplot(Nd ~ Type, xlab="", xaxt="n", cex.lab=1.8, 
     ylab=expression(bold(paste(Nd[sample],"/Nd"[PAAS]))), ylim=c(0,3.8), 
     col=c("royalblue","wheat","honeydew3","lightblue"), cex=1.5))
axis(1, at=1:4, labels=c("Drain\nsediment","Lake\nsediment","Saltmarsh\n",
        "Flooded\nwoodland"), mgp=c(3,2,0), gap.axis=0)
abline(h=0.98, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
with(afs1922_PAAS, boxplot(Gd~Type, xlab="",xaxt="n", cex.lab=1.8, ylim=c(0,3.8), 
     ylab=expression(bold(paste(Gd[sample],"/Gd"[PAAS]))), col=c("royalblue","wheat","honeydew3","lightblue"), cex=1.5))
axis(1, at=1:4, labels=c("Drain\nsediment","Lake\nsediment","Saltmarsh\n",
        "Flooded\nwoodland"), mgp=c(3,2,0), gap.axis=0)
abline(h=0.98, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
with(afs1922_PAAS, boxplot(Y ~ Type, xlab="", xaxt="n", cex.lab=1.8, 
     ylab=expression(bold(paste(Y[sample],"/Y"[PAAS]))), ylim=c(0,3.8), 
     col=c("royalblue","wheat","honeydew3","lightblue"), cex=1.5))
axis(1, at=1:4, labels=c("Drain\nsediment","Lake\nsediment","Saltmarsh\n",
        "Flooded\nwoodland"), mgp=c(3,2,0), gap.axis=0)
abline(h=0.98, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
```

### ├ mean comparisons and pairwise tests

```{r load-packs-}
library(multcompView)
library(rcompanion)
```

```{r REE-by-type, results='hold'}
with(afs1922_PAAS, signif(tapply(REE, Type, mean, na.rm=TRUE),3))
with(afs1922_PAAS, signif(tapply(REE, Type, sd, na.rm=TRUE),3))
with(afs1922_PAAS, oneway.test(REE ~ Type))
(pw0 <- with(afs1922_PAAS, pairwise.t.test(REE, Type)))
fpw0 <- fullPTable(pw0$p.value)
(multcompLetters(fpw0))
```

```{r}
with(afs1922_PAAS, signif(tapply(La, Type, mean, na.rm=TRUE),2))
with(afs1922_PAAS, signif(tapply(La, Type, sd, na.rm=TRUE),2))
with(afs1922_PAAS, oneway.test(La ~ Type))
(pw0 <- with(afs1922_PAAS, pairwise.t.test(La, Type)))
fpw0 <- fullPTable(pw0$p.value)
(multcompLetters(fpw0))
```

```{r}
with(afs1922_PAAS, signif(tapply(Ce, Type, mean, na.rm=TRUE),2))
with(afs1922_PAAS, signif(tapply(Ce, Type, sd, na.rm=TRUE),2))
with(afs1922_PAAS, oneway.test(Ce ~ Type))
(pw0 <- with(afs1922_PAAS, pairwise.t.test(Ce, Type)))
fpw0 <- fullPTable(pw0$p.value)
(multcompLetters(fpw0))
```

```{r}
with(afs1922_PAAS, signif(tapply(Nd, Type, mean, na.rm=TRUE),2))
with(afs1922_PAAS, signif(tapply(Nd, Type, sd, na.rm=TRUE),2))
with(afs1922_PAAS, oneway.test(Nd ~ Type))
(pw0 <- with(afs1922_PAAS, pairwise.t.test(Nd, Type)))
fpw0 <- fullPTable(pw0$p.value)
(multcompLetters(fpw0))
```

```{r}
with(afs1922_PAAS, signif(tapply(Gd, Type, mean, na.rm=TRUE),2))
with(afs1922_PAAS, signif(tapply(Gd, Type, sd, na.rm=TRUE),2))
with(afs1922_PAAS, oneway.test(Gd ~ Type))
(pw0 <- with(afs1922_PAAS, pairwise.t.test(Gd, Type)))
fpw0 <- fullPTable(pw0$p.value)
(multcompLetters(fpw0))
```

```{r}
with(afs1922_PAAS, signif(tapply(Y, Type, mean, na.rm=TRUE),2))
with(afs1922_PAAS, signif(tapply(Y, Type, sd, na.rm=TRUE),2))
with(afs1922_PAAS, oneway.test(Y ~ Type))
(pw0 <- with(afs1922_PAAS, pairwise.t.test(Y, Type)))
fpw0 <- fullPTable(pw0$p.value)
(multcompLetters(fpw0))
```

# By individual sample Type

## ├ YREE/PAAS in drain sediments

```{r normalized-REE-Drain_sed, fig.height=3.6, fig.width=6, results='hold'}
PAAS_DrainSed <- subset(afs1922_PAAS, afs1922_PAAS$Type=="Drain_Sed")
plotdata <- 
  stack(st_drop_geometry(PAAS_DrainSed[,c("La","Ce","Nd","Gd","Y","REE")]))
par(mfrow=c(1,1), mar = c(2,4,1,1), mgp=c(1.7,0.3,0), tcl=0.25, font.lab=2,
    lend=2, ljoin=1)
meanz <- with(plotdata, tapply(values, ind, FUN=mean, na.rm=TRUE))
with(plotdata, boxplot(values ~ ind, xaxt="n", xlab="", 
              ylab="Normalized REE = sample/PAAS", col = inferno(9, dir=1)[4:9]))
points(c(1:6), as.numeric(meanz), pch=3, col="white", lwd=4)
points(c(1:6), as.numeric(meanz), pch=3, lwd=2)
axis(1, at=1:6, labels=c("La","Ce","Nd","Gd","Y","∑REE"), font=2)
abline(h=0.99, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
mtext("Drain Sediment", side=3, line=-1.2, adj=0.98, font=2, cex = 1.2)
```
```{r}
with(plotdata, oneway.test(values ~ ind))
with(plotdata, pairwise.t.test(values, ind))
```

Drain sediments are depleted in REE relative to PAAS. Gd/PAAS is greatest but not different from La, Ce or Nd, only greater than Y/PAAS (lowest mean).

## ├ YREE/PAAS in Lake sediments

```{r normalized-REE-Lake_sed, fig.height=3.6, fig.width=6, results='hold'}
require(viridis)
PAAS_LakeSed <- subset(afs1922_PAAS, afs1922_PAAS$Type=="Lake_Sed")
plotdata <- 
  stack(st_drop_geometry(PAAS_LakeSed[,c("La","Ce","Nd","Gd","Y","REE")]))
par(mfrow=c(1,1), mar = c(2,4,1,1), mgp=c(1.7,0.3,0), tcl=0.25, font.lab=2,
    lend=2, ljoin=1)
meanz <- with(plotdata, tapply(values, ind, FUN=mean, na.rm=TRUE))
with(plotdata, boxplot(values ~ ind, xaxt="n", xlab="", 
              ylab="Normalized REE = sample/PAAS", col = plasma(9, dir=1)[4:9]))
points(c(1:6), as.numeric(meanz), pch=3, col="white", lwd=4)
points(c(1:6), as.numeric(meanz), pch=3, lwd=2)
axis(1, at=1:6, labels=c("La","Ce","Nd","Gd","Y","∑REE"), font=2)
abline(h=0.99, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
mtext("Lake Sediment", side=3, line=-1.2, adj=0.98, font=2, cex = 1.2)
```

Wetland pond ("Lake") sediments are enriched relative to PAAS. Y is significantly less than La, Ce, Nd, and Gd.

```{r}
with(plotdata, oneway.test(values ~ ind))
with(plotdata, pairwise.t.test(values, ind))
```

## ├ YREE/PAAS in Saltmarsh sediments

```{r normalized-REE-Drain_sed, fig.height=3.6, fig.width=6, results='hold'}
require(viridis)
PAAS_SM <- subset(afs1922_PAAS, afs1922_PAAS$Type=="Saltmarsh")
plotdata <- 
  stack(st_drop_geometry(PAAS_SM[,c("La","Ce","Nd","Gd","Y","REE")]))
par(mfrow=c(1,1), mar = c(2,4,1,1), mgp=c(1.7,0.3,0), tcl=0.25, font.lab=2,
    lend=2, ljoin=1)
meanz <- with(plotdata, tapply(values, ind, FUN=mean, na.rm=TRUE))
with(plotdata, boxplot(values ~ ind, xaxt="n", xlab="", 
              ylab="Normalized REE = sample/PAAS", col = viridis(9, dir=1)[4:9]))
points(c(1:6), as.numeric(meanz), pch=3, col="white", lwd=4)
points(c(1:6), as.numeric(meanz), pch=3, lwd=2)
axis(1, at=1:6, labels=c("La","Ce","Nd","Gd","Y","∑REE"), font=2)
abline(h=0.99, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
mtext("Saltmarsh Sediment", side=3, line=-1.2, adj=0.98, font=2, cex = 1.2)
```
```{r}
with(plotdata, oneway.test(values ~ ind))
with(plotdata, pairwise.t.test(values, ind))
```

Saltmarsh sediments are enriched in YREE relative to PAAS. Gd > all others
except La. Y < all others.

## ├ YREE/PAAS in Flooded woodland sediments

```{r normalized-REE-Drain_sed, fig.height=3.6, fig.width=6, results='hold'}
require(viridis)
PAAS_Flooded <- subset(afs1922_PAAS, afs1922_PAAS$Type=="Flooded")
plotdata <- 
  stack(st_drop_geometry(PAAS_Flooded[,c("La","Ce","Nd","Gd","Y","REE")]))
par(mfrow=c(1,1), mar = c(2,4,1,1), mgp=c(1.7,0.3,0), tcl=0.25, font.lab=2,
    lend=2, ljoin=1)
meanz <- with(plotdata, tapply(values, ind, FUN=mean, na.rm=TRUE))
with(plotdata, boxplot(values ~ ind, xaxt="n", xlab="", 
              ylab="Normalized REE = sample/PAAS", col = mako(9, dir=1)[4:9]))
points(c(1:6), as.numeric(meanz), pch=3, col="white", lwd=4)
points(c(1:6), as.numeric(meanz), pch=3, lwd=2)
axis(1, at=1:6, labels=c("La","Ce","Nd","Gd","Y","∑REE"), font=2)
abline(h=0.995, lty=3,col="sienna4");abline(h=1, lty=3, col="gold")
mtext("Flooded Woodland", side=3, line=-1.2, adj=0.75, font=2, cex = 1.2)
```

```{r}
with(plotdata, oneway.test(values ~ ind))
with(plotdata, pairwise.t.test(values, ind))
```

```{r}

```

```{r tidy up, message=FALSE, warning=FALSE, include=FALSE}
rm(list = c("s0","ns0","d0","f0","p1","p2","varz","clrz","x0","x1","lm0","v0",
            "data0","mcl0"))
```

#### end of file
