---
title: "Ashfield 2021-2023 XRD summaries and PCA"
output: 
  bookdown::html_document2:
    fig_width: 6.4
    fig_height: 4.8
    number_sections: no
    self_contained: no
    df_print: default
    fig_caption: yes
    smart: no
---

```{r load packages and environment but do not show, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}
library(mgsub)
library(stringr)
library(flextable)
  set_flextable_defaults(font.family = 'sans', theme_fun = "theme_zebra", 
        font.size = 11, text.align = "right",padding.left = 1, padding.right = 1,
        padding.top = 2, padding.bottom = 2)
library(lctools)
library(gstat)
library(sf)
library(maptiles)
library(prettymapr)
library(viridis)
library(scico)
```

```{r more-setup, message=FALSE, warning=FALSE, echo=FALSE}
git <- "https://github.com/Ratey-AtUWA/Ashfield-REE/raw/main/"
afr_map <- read.csv(file=paste0(git,"afr_map_v3.csv"), stringsAsFactors = TRUE)
pal4lite <- c("black", plasma(8), "white")
pal4liteTransp <- 
  c("black", scico(8, alpha=0.7, palette="hawaii", beg=0.1), "white")
pal4dark <- c("white", turbo(8, beg=0.2, end=0.9,dir=1), "black")
```

Sediment samples were homogenised, and a 20 g subsample was taken for
preparation for XRD analysis. Organic matter was removed from samples by
suspending samples in 30% H~2~O~2~ with gentle heating, and the organic-free samples
were dried at 60&nbsp;°C.

Sub-samples for random powder XRD analysis were prepared from the organic-free
material by The clay fraction of sediment samples was separated by sedimentation
after dispersion in 0.01M sodium hexametaphosphate solution. Using Stokes Law to
estimate sampling time and depth, the suspension containing only particles
≤&nbsp;2&nbsp;µm was collected on a porous sample holder to provide an oriented
sample for XRD analysis. After initial saturation with MgCl~2~, subsequent
washing with H~2~O (to remove excess MgCl~2~), and air drying, powder
diffraction patterns were acquired using a Panalytical Aeris X-ray
Diffractometer. The same sample was similarly treated with 50% glycerol
solution, and the diffraction pattern remeasured. Samples suspected to contain
vermiculite and/or interstratified 2:1 phyllosilicates were heated at
400&nbsp;°C and 550&nbsp;°C and remeasured after each heating step. Clay X-ray
diffraction patterns were collected using Co-Kα1 radiation generated at
40&nbsp;kV and 15&nbsp;mA using a Ni filter, with a goniometer range of
3-30&nbsp;°2θ and step size of 0.00109&nbsp;°2θ, and a total scan time of 8
minutes. XRD patterns were analysed using High Score Plus software (Degen, 2014)
in conjunction with the USGS clay mineral identification key (Poppe, 2004).

<p>&nbsp;</p>

# Summary of random powder XRD data

```{r clay-table, results='hold', echo=FALSE}
afxrd2123 <- read.csv("afxrd2123corr.csv", stringsAsFactors = T)
afxrd2123$Sample <- as.character(afxrd2123$Sample)

flextable(afxrd2123) |>
  rotate(j=6:38, rotation="btlr", part="header") |>
  valign(valign="bottom", part="header") |>
  set_caption(caption = "Summary of random powder XRD for Ashfield Flats samples 2021-2023. Values are percentages based on semiquantitative Rietveld refinement of XRD patterns. Feldspars (mainly albite, microcline, orthoclase, and anorthite) have been combined into a single category.") |>
  font(fontname = "Arial Narrow", part="all") |>
  vline(border=officer::fp_border(color = "grey65", style = "solid", width = 0.5))
```

<p>&nbsp;</p>

```{r map-pyrite-jarosite, fig.width=10.4, fig.height=7, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Locations at Ashfield Flats where pyrite and jarosite were detected by XRD in sediment samples.", results='hold'}
extent <- st_as_sf(x = data.frame(x = c(399880,400520),
                                  y = c(6467960,6468360)),
                   coords = c("x", "y"), crs = st_crs(32750))
aftiles <- get_tiles(extent, provider="Thunderforest.OpenCycleMap",
             apikey="03b0375e8e7c47d0a03018a9001de439",
                      crop = TRUE, zoom = 17) # make map object
par(mfrow=c(1,1), mar=c(0,0,0,0), oma=c(3,3,0.5,1.5), mgp=c(1.6,0.2,0),
    tcl=0.3,lend="square", ljoin="mitre")
plot_tiles(aftiles)
axis(1);axis(2);box()
mtext(side=1, line=1.6, text="Easting (UTM Zone 50, m)", font=2, cex=1)
mtext(side=2, line=1.6, text="Northing (UTM Zone 50, m)", font=2, cex=0.9)
addnortharrow(pos="topleft", border=1, lwd=1, text.col=1,
padin=c(0.05,0.05), scale = 0.7)
addscalebar(plotepsg = 32750, label.cex = 1.2)
with(afr_map, polygon(wetland_E, wetland_N, col="skyblue",
border="steelblue"))
with(afr_map, lines(drain_E, drain_N, col="steelblue"))
with(subset(afxrd2123,afxrd2123$Type!="Core"),
     points(Easting, Northing, cex=sqrt(Jarosite),
                       pch=22, bg="#fff08080"))
with(subset(afxrd2123,afxrd2123$Type!="Core"),
     points(Easting, Northing, cex=sqrt(Pyrite),
            pch=21, bg="#8000c080"))
legend("bottomright", legend = c("10%","1%",NA,"Pyrite","Jarosite"),
       bty="n", inset=0.02, title=expression(bold("Mineral content")),
       pt.cex=c(3.16,1,NA,2,2), pch=c(21,21,NA,21,22),
       pt.bg=c("#80808080","#80808080",NA,"#8000c080","#fff08080"))

```

```{r map-calcite-gypsum, fig.width=10.4, fig.height=7, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Locations at Ashfield Flats where calcite and gypsum were detected by XRD in sediment samples.", echo=FALSE, results='hold'}
# calcite & gypsum
par(mfrow=c(1,1), mar=c(0,0,0,0), oma=c(3,3,0.5,1.5), mgp=c(1.6,0.2,0),
    tcl=0.3,lend="square", ljoin="mitre")
plot_tiles(aftiles)
axis(1);axis(2);box()
mtext(side=1, line=1.6, text="Easting (UTM Zone 50, m)", font=2, cex=1)
mtext(side=2, line=1.6, text="Northing (UTM Zone 50, m)", font=2, cex=0.9)
addnortharrow(pos="topleft", border=1, lwd=1, text.col=1,
              padin=c(0.05,0.05), scale = 0.7)
addscalebar(plotepsg = 32750, label.cex = 1.2)
with(afr_map, polygon(wetland_E, wetland_N, col="skyblue",
                      border="steelblue"))
with(afr_map, lines(drain_E, drain_N, col="steelblue"))
with(subset(afxrd2123,afxrd2123$Type!="Core"),
     points(Easting, Northing, cex=sqrt(Calcite),
            pch=21, bg="#ffffff80"))
with(subset(afxrd2123,afxrd2123$Type!="Core"),
     points(Easting, Northing, cex=sqrt(Gypsum),
            pch=22, bg="#ff808080"))
legend("bottomright", legend = c("10%","1%",NA,"Calcite","Gypsum"),
       bty="n", inset=0.02, title=expression(bold("Mineral content")),
       pt.cex=c(3.16,1,NA,2,2), pch=c(21,21,NA,21,22),
       pt.bg=c("#b0b0b080","#b0b0b080",NA,"#ffffff80","#ff808080"))

```

# Pricipal Components Analysis using mineral percentages

```{r clr-transform-mineral-data, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}
afxrd_clr <- afxrd2123
afxrd_clr[,8:38] <- t(apply(afxrd_clr[,8:38], MARGIN = 1,
                            FUN = function(x){log(x) - mean(log(x),na.rm=T)}))
# replace NAs with nominal (low) value
for(i in 8:38){
  afxrd_clr[which(is.na(afxrd_clr[,i])),i] <- -7
}
```

```{r do-pca, message=FALSE, warning=FALSE, results='hold'}
af_min_pca <- prcomp(afxrd_clr[,c(8:38)], scale. = TRUE)
summary(af_min_pca)
```

<p>&nbsp;</p>

```{r show-variables-influencing-PCs, message=FALSE, warning=FALSE, paged.print=FALSE, echo=FALSE, results='hold'}
x0 <- 1      # ■■■ CHOOSE THE PRINCIPAL COMPONENT TO BE MAPPED ■■■
varwts1 <- data.frame(PC1.var=row.names(af_min_pca$rotation),
                      PC1=round(af_min_pca$rotation[,x0],3),
                      PC1abs=abs(af_min_pca$rotation[,x0]))
varwts1 <- varwts1[rev(order(varwts1$PC1abs)),] ; row.names(varwts1) <- NULL
x0 <- 2      # ■■■ CHOOSE THE PRINCIPAL COMPONENT TO BE MAPPED ■■■
varwts2 <- data.frame(PC2.var=row.names(af_min_pca$rotation),
                      PC2=round(af_min_pca$rotation[,x0],3),
                      PC2abs=abs(af_min_pca$rotation[,x0]))
varwts2 <- varwts2[rev(order(varwts2$PC2abs)),] ; row.names(varwts2) <- NULL
x0 <- 3      # ■■■ CHOOSE THE PRINCIPAL COMPONENT TO BE MAPPED ■■■
varwts3 <- data.frame(PC3.var=row.names(af_min_pca$rotation),
                      PC3=round(af_min_pca$rotation[,x0],3),
                      PC3abs=abs(af_min_pca$rotation[,x0]))
varwts3 <- varwts3[rev(order(varwts3$PC3abs)),] ; row.names(varwts3) <- NULL
x0 <- 4      # ■■■ CHOOSE THE PRINCIPAL COMPONENT TO BE MAPPED ■■■
varwts4 <- data.frame(PC4.var=row.names(af_min_pca$rotation),
                      PC4=round(af_min_pca$rotation[,x0],3),
                      PC4abs=abs(af_min_pca$rotation[,x0]))
varwts4 <- varwts4[rev(order(varwts4$PC4abs)),] ; row.names(varwts4) <- NULL
summwts <- cbind(varwts1[,1:2],varwts2[,1:2],varwts3[,1:2],varwts4[,1:2])
flextable(summwts[1:8,]) |>
  set_caption(caption="Minerals having the greatest absolute weightings for each principal component for the Ashfield Flats sediment data. Feldspars (mainly albite, microcline, orthoclase, and anorthite) have been combined into a single category.") |>
  flextable::set_header_labels(values=list(PC1.var="Mineral", 
                                           PC1="PC1 weighting", 
                                           PC2.var="Mineral", 
                                           PC2="PC2 weighting", 
                                           PC3.var="Mineral", 
                                           PC3="PC3 weighting", 
                                           PC4.var="Mineral", 
                                           PC4="PC4 weighting")) |>
  vline(j=c(2,4,6), border=officer::fp_border(color = "grey65", style = "solid", width = 0.5)) |>
  valign(valign="bottom", part="header") |>
  padding(j=c(1,3,5,7), padding.left=8, part="all") |>
  padding(j=c(2,4,6,8), padding.right=8, part="all") |>
  width(j=c(2,4,6,8), width=2.5, unit="cm") |>
  width(j=c(1,3,5,7), width=3.5, unit="cm")
```

<p>&nbsp;</p>

## PCA biplot

```{r xrd-pca-biplot, fig.width=14, fig.height=7, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Biplots summarising principal components analysis for centered-logratio transformed mineral contents in Ashfield Flats sediments. Feldspars (mainly albite, microcline, orthoclase, and anorthite) have been combined into a single category.", results='hold'}
par(mfrow=c(1,2), oma=c(0,0,0,0), mar=c(3.5,3.5,3,3), mgp=c(1.7,0.3,0), tcl=0.25, 
    font.lab=2, font.axis=1)
biplot(af_min_pca, choices = c(1,2), col=c("#00000000","grey35"),
       scale=.4, xlab="PC1", ylab="PC2", cex.lab=1.4, font.lab=2,
       xlim=c(-0.8,2.3), ylim=c(-1,1.1)
       )
ff <- 0.25
with(af_min_pca, points(x[,1]*ff, x[,2]*ff, cex=1.75,
                        pch=c(21,22,23,24,25,21,22,23,24)[afxrd_clr$Zone],
                        bg=plasma(9, alpha=0.7)[afxrd_clr$Zone]))
legend("topright", bty="n", inset=0.005, cex=1.2, ncol=2, y.int=0.85,
       title=expression(bold("Ashfield Zone")),
       legend=levels(afxrd_clr$Zone),
       pch=c(21,22,23,24,25,21,22,23,24), pt.cex=1.5,
       pt.bg=plasma(9,alp=0.7))

biplot(af_min_pca, choices = c(1,2), col=c("#00000000","grey35"),
       scale=.7, xlab="PC1", ylab="PC2", cex.lab=1.4, font.lab=2,
       xlim=c(-0.4,0.4), ylim=c(-0.2,0.32)
       )
ff <- 1.4
with(af_min_pca, points(x[,1]*ff, x[,2]*ff, cex=1.75,
                        pch=c(21,22,23,24,25,21,22,23,24)[afxrd_clr$Zone],
                        bg=plasma(9, alpha=0.7)[afxrd_clr$Zone]))
mtext("Zoomed in to show central\ncluster of observations",1,-1.5,adj=0.02, font=2)
```

The north pond, southwest saltmarsh, and some samples from the main saltmarsh
have positive observation scores for both PC1 and PC2, consistent with greater
pyrite and jarosite (PC1) and greater kaolinite and illite (*i.e*. clay) (PC2)
(Table \@ref(tab:show-variables-influencing-PCs), Figure \@ref(fig:xrd-pca-biplot)).

----------------------

## Pricipal Components Analysis using only more common mineral percentages

```{r clr-transform-reduced-mineral-data, message=FALSE, warning=FALSE, paged.print=FALSE, results='hold'}
afxrd_clr_min <- afxrd2123[,c("Easting","Northing","Type","Zone",
                          "Quartz","Kaolinite","Illite","Feldspars","Halite",
                          "Gypsum","Pyrite","Jarosite","Hornblende","Calcite","Rutile")]
afxrd_clr_min[,5:15] <- t(apply(afxrd_clr_min[,5:15], MARGIN = 1,
                            FUN = function(x){log(x) - mean(log(x),na.rm=T)}))
# replace NAs with nominal (low) value
for(i in 5:15){
  afxrd_clr_min[which(is.na(afxrd_clr_min[,i])),i] <- -7
}
```

```{r do-pca-reduced-data, message=FALSE, warning=FALSE, results='hold'}
af_common_pca <- prcomp(afxrd_clr_min[,c(5:15)], scale. = TRUE)
summary(af_common_pca)
```

<p>&nbsp;</p>

```{r reduced-variables-influencing-PCs, message=FALSE, warning=FALSE, paged.print=FALSE, echo=FALSE, results='hold'}
x0 <- 1      # ■■■ CHOOSE THE PRINCIPAL COMPONENT TO BE MAPPED ■■■
varwts1 <- data.frame(PC1.var=row.names(af_common_pca$rotation),
                      PC1=round(af_common_pca$rotation[,x0],3),
                      PC1abs=abs(af_common_pca$rotation[,x0]))
varwts1 <- varwts1[rev(order(varwts1$PC1abs)),] ; row.names(varwts1) <- NULL
x0 <- 2      # ■■■ CHOOSE THE PRINCIPAL COMPONENT TO BE MAPPED ■■■
varwts2 <- data.frame(PC2.var=row.names(af_common_pca$rotation),
                      PC2=round(af_common_pca$rotation[,x0],3),
                      PC2abs=abs(af_common_pca$rotation[,x0]))
varwts2 <- varwts2[rev(order(varwts2$PC2abs)),] ; row.names(varwts2) <- NULL
x0 <- 3      # ■■■ CHOOSE THE PRINCIPAL COMPONENT TO BE MAPPED ■■■
varwts3 <- data.frame(PC3.var=row.names(af_common_pca$rotation),
                      PC3=round(af_common_pca$rotation[,x0],3),
                      PC3abs=abs(af_common_pca$rotation[,x0]))
varwts3 <- varwts3[rev(order(varwts3$PC3abs)),] ; row.names(varwts3) <- NULL
x0 <- 4      # ■■■ CHOOSE THE PRINCIPAL COMPONENT TO BE MAPPED ■■■
varwts4 <- data.frame(PC4.var=row.names(af_common_pca$rotation),
                      PC4=round(af_common_pca$rotation[,x0],3),
                      PC4abs=abs(af_common_pca$rotation[,x0]))
varwts4 <- varwts4[rev(order(varwts4$PC4abs)),] ; row.names(varwts4) <- NULL
summwts <- cbind(varwts1[,1:2],varwts2[,1:2],varwts3[,1:2],varwts4[,1:2])
flextable(summwts[1:5,]) |>
  set_caption(caption="Minerals having the greatest absolute weightings for each principal component for the Ashfield Flats sediment data. Feldspars (mainly albite, microcline, orthoclase, and anorthite) have been combined into a single category.") |>
  flextable::set_header_labels(values=list(PC1.var="Mineral", 
                                           PC1="PC1 weighting", 
                                           PC2.var="Mineral", 
                                           PC2="PC2 weighting", 
                                           PC3.var="Mineral", 
                                           PC3="PC3 weighting", 
                                           PC4.var="Mineral", 
                                           PC4="PC4 weighting")) |>
  vline(j=c(2,4,6), border=officer::fp_border(color = "grey65", style = "solid", width = 0.5)) |>
  valign(valign="bottom", part="header") |>
  padding(j=c(1,3,5,7), padding.left=8, part="all") |>
  padding(j=c(2,4,6,8), padding.right=8, part="all") |>
  width(j=c(2,4,6,8), width=2.5, unit="cm") |>
  width(j=c(1,3,5,7), width=3.5, unit="cm")
```

<p>&nbsp;</p>

## PCA 1-2 biplot

```{r xrd-pca-biplot-reduced, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Biplots summarising principal components analysis for centered-logratio transformed proportions of the more common mineral phases in Ashfield Flats sediments. Feldspars (mainly albite, microcline, orthoclase, and anorthite) have been combined into a single category.", results='hold'}
par(mfrow=c(1,2), oma=c(0,0,0,0), mar=c(3.5,3.5,1.5,0.5), mgp=c(1.7,0.3,0), tcl=0.25, 
    font.lab=2, font.axis=1)
biplot(af_common_pca, choices = c(1,2), col=c("#00000000","grey35"),
       scale=.4, xlab="PC1", ylab="PC2", cex.lab=1.4, font.lab=2,
       xlim=c(-0.55,0.8), ylim=c(-0.6,0.75)
       )
ff <- 0.24
with(af_common_pca, points(x[,1]*ff, x[,2]*ff, cex=1.75,
                        pch=c(21,22,23,24,25,21,22,23,24)[afxrd_clr_min$Zone],
                        bg=plasma(9, alpha=0.7)[afxrd_clr_min$Zone]))

plot(0:1,0:1, bty="n",axes=F,xlab="",ylab="",type="n")
legend("left", bty="n", inset=0.005, cex=1.2, ncol=1, y.int=1,
       title=expression(bold("Ashfield Zone")),
       legend=levels(afxrd_clr_min$Zone),
       pch=c(21,22,23,24,25,21,22,23,24), pt.cex=1.5,
       pt.bg=plasma(9,alp=0.7))
```

```{r xrd-pca-biplot-34-reduced, fig.width=12, fig.height=6, message=FALSE, warning=FALSE, fig.align='center', fig.cap="Biplots summarising principal components analysis for centered-logratio transformed proportions of the more common mineral phases in Ashfield Flats sediments. Feldspars (mainly albite, microcline, orthoclase, and anorthite) have been combined into a single category.", results='hold'}
par(mfrow=c(1,2), oma=c(0,0,0,0), mar=c(3.5,3.5,1.5,0.5), mgp=c(1.7,0.3,0), tcl=0.25, 
    font.lab=2, font.axis=1)
biplot(af_common_pca, choices = c(3,4), col=c("#00000000","grey35"),
       scale=.4, xlab="PC3", ylab="PC4", cex.lab=1.4, font.lab=2,
       #xlim=c(-0.55,0.8), ylim=c(-0.6,0.75)
       )
ff <- 0.4
with(af_common_pca, points(x[,3]*ff, x[,4]*ff, cex=1.75,
                        pch=c(21,22,23,24,25,21,22,23,24)[afxrd_clr_min$Zone],
                        bg=plasma(9, alpha=0.7)[afxrd_clr_min$Zone]))

plot(0:1,0:1, bty="n",axes=F,xlab="",ylab="",type="n")
legend("left", bty="n", inset=0.005, cex=1.2, ncol=1, y.int=1,
       title=expression(bold("Ashfield Zone")),
       legend=levels(afxrd_clr_min$Zone),
       pch=c(21,22,23,24,25,21,22,23,24), pt.cex=1.5,
       pt.bg=plasma(9,alp=0.7))
```

-------------------------

That's it so far
